---
title: 'Residency and exhaustion signatures for CD4 T cells in Zhang et al Smart-seq2 single cell data'
author: 
  - name: 'Sepideh Foroutan'
    affiliation: 'Huntington Cancer Immunotherapy Lab'
    url: https://www.monash.edu/discovery-institute/huntington-lab
date: '10-06-2020'
output:
  rmdformats::readthedown:
    fig_width: 12
    fig_height: 6
    gallery: TRUE
    highlight: tango
    lightbox: TRUE
    self_contained: TRUE
    thumbnails: FALSE
    number_sections: TRUE	
    toc_depth: 3
    use_bookdown: TRUE
    code_folding: hide
  html_document2:
    df_print: paged
params:
  update_date: !r paste("Last updated on:", Sys.Date() )
editor_options: 
  chunk_output_type: inline
---
`r params$update_date`

<style>
body {
text-align: justify}
</style>


```{r knitr_init, echo=FALSE, results="asis", cache=FALSE}

library(knitr)
library(rmdformats)

options(max.print = "75")
opts_chunk$set(
  # echo = FALSE,
  cache = TRUE,
  # prompt = FALSE,
  # tidy = FALSE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  # results = FALSE,
  root.dir = normalizePath("."))
opts_knit$set(width = 65)

```

# Set up and overview
```{r file set up, echo=F, results=F, message=F, warning=F}

library(DT)
library(Seurat)
library(SingleCellExperiment)
library(singscore)
library(slingshot)
library(tidyverse)
library(RColorBrewer)

mainDir <- getwd()
outPath <- "../output/CD4/"
figPath <- "../figure/CD4/"
dataPath <- "../data/"
scriptPath <- "../script/"


tpmPath <- "../data/GSE146771_Zhang_CRC/"
hpaGoUniprotPath <- "../data/HPA_GO_Uniprot/"
ccleLcmPath <- "../data/CCLE_LCM/"

ifelse(!dir.exists(file.path(mainDir, outPath)), dir.create(file.path(mainDir, outPath)), FALSE)
ifelse(!dir.exists(file.path(mainDir, figPath)), dir.create(file.path(mainDir, figPath)), FALSE)
ifelse(!dir.exists(file.path(mainDir, dataPath)), dir.create(file.path(mainDir, dataPath)), FALSE)
ifelse(!dir.exists(file.path(mainDir, scriptPath)), dir.create(file.path(mainDir, scriptPath)), FALSE)


options(digits = 3)

equal_breaks <- function(n = nBreak, s = scalingFactor, ...){
  function(x){
    # rescaling
    d <- s * diff(range(x)) / (1+2*s)
    round( seq(min(x)+d, max(x)-d, length=n), 2)
  }
}

nBreak = 3
scalingFactor = 0.05

textSize <- 1.2

currentTheme <- theme_minimal() +
  theme(
    # panel.background = element_blank()
    axis.title = element_text(size = rel(textSize)),
    axis.text = element_text(angle = 0, size = rel(textSize)),
    # strip.background = element_rect(colour = "#f0f0f0", fill = "#f0f0f0"),
    strip.text = element_text(size = rel(textSize)),
    axis.line = element_line(colour = "black", size = 0.5),
    legend.position = 'top',
    # legend.title = element_text(size = rel(textSize), face = "italic"),
    # legend.text = element_text(size = rel(textSize)),
    legend.key.size = unit(1, 'lines'),
    ## increase the line space
    plot.title = element_text(
      face = "bold",
      size = rel(textSize),
      hjust = 0.5
    )
  )

cols <-  c(
  brewer.pal(8, "Dark2")[-5],
  brewer.pal(10, "Paired"),
  brewer.pal(12, "Set3"),
  brewer.pal(9, "Blues")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Oranges")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Greens")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Purples")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Reds")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Greys")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "BuGn")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "PuRd")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "BuPu")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "YlGn")[c(8, 3, 7, 4, 6, 9, 5)])

source(paste0(scriptPath, "visReduction_function.R"))
```


# Read SMART-seq2 CD4 data
We read the SMART-seq2 data, and generate a Seurat object for it. The data is already in TPM format.
Note that among 18 patients in Zhang et al study, 3 of them are MSI-H (P0123, P0413 and P0825), and the other 15 patients are MSS. So we add these info to the annotation file and calculate the percent of mitochondrial genes.
```{r, cache = T}
## dim:15179 10469

crcSmart <-
   data.table::fread(
    paste0(tpmPath, "GSE146771_CRC.Leukocyte.Smart-seq2.TPM.txt")
  )

rnames <- crcSmart$V1
crcSmart <- crcSmart[, 2:ncol(crcSmart)]
rownames(crcSmart) <- rnames

crcSmart <- as.matrix(crcSmart)
rownames(crcSmart) <- rnames


metaSmart <-
  read.table(
    paste0(tpmPath, "GSE146771_CRC.Leukocyte.Smart-seq2.Metadata.txt"),
    header = T,
    sep = "\t"
  )

metaSmart$Sub_Cluster <- as.character(metaSmart$Sub_Cluster)
metaSmart$Sub_Cluster <- sapply(metaSmart$Sub_Cluster, function(x){
  unlist(strsplit(x, "_"))[2]
})


##----- assign colours to cells
immCells <- metaSmart$Sub_Cluster[grepl("CD8", metaSmart$Sub_Cluster)|
                                    grepl("NK", metaSmart$Sub_Cluster) |
                                    grepl("CD4", metaSmart$Sub_Cluster)|
                                    grepl("ILC3", metaSmart$Sub_Cluster)]

cellCols <- cols[1:length(unique(immCells))]
names(cellCols) <- unique(immCells)[order(unique(immCells))]
cd4Cols <- cellCols[grepl("CD4", 
                          names(cellCols))]

##----- generate Seurat object

subSmartCD4 <- crcSmart[, grepl('CD4', metaSmart$Sub_Cluster) ]
subMetaSmartCD4 <- metaSmart[grepl('CD4', metaSmart$Sub_Cluster), ]


# all(subMetaSmartCD4$CellName == colnames(subSmartCD4))
rownames(subMetaSmartCD4) <- subMetaSmartCD4$CellName

subMetaSmartCD4$Sub_Cluster <- as.character(subMetaSmartCD4$Sub_Cluster)

sdRow <- matrixStats::rowSds(subSmartCD4)
subSmartCD4 <- subSmartCD4[ sdRow!=0 , ]

cd4 <-
  CreateSeuratObject(
    counts = subSmartCD4,
    project = 'Zhang_SmartSeq_CD4',
    meta.data = subMetaSmartCD4,
    min.cells = 5
  )

cd4[['percent.mt']] <- PercentageFeatureSet(cd4, pattern = '^MT-')

cd4$MSI <- "MSS"
cd4$MSI[cd4$Sample %in% c("P0123", "P0413", "P0825")] <- "MSI-H"

## save the RDS file of the CD4
# saveRDS(cd4, paste0(dataPath, "GSE146771_Zhang_CRC/CD4_Zhang_Smartseq_Seurat.RDS"))
```


## Read GPCRs and TFs
In order to annotate genes based on whether or not they are TFs, GPCR or associated with GPCR signalling, we use the human protein atlas and get the GPCR and TF annotations from there. For GPCR, we also downloaded the GPCR related genes from QuickGO and selected the genes annotated as gpcr in uniprot too. 
```{r}
hpa <- read.delim(paste0(hpaGoUniprotPath, "proteinatlas.tsv"), header = T)

GPCRs <-
  hpa[grepl("coupled", hpa$Gene.description, ignore.case = T) |
        grepl("coupled", hpa$Protein.class, ignore.case = T), c(
          "Gene",
          "Ensembl",
          "Protein.class",
          "Gene.description",
          "Uniprot" ,
          "Blood.RNA...NK.cell..NX."
        )]

GPCRs <- GPCRs[order(GPCRs$Blood.RNA...NK.cell..NX., decreasing = T), ]


TFs <- hpa[grepl("Transcription factor", hpa$Gene.description, ignore.case = T) |
        grepl("Transcription factor", hpa$Protein.class, ignore.case = T), c(
          "Gene",
          "Ensembl",
          "Protein.class",
          "Gene.description",
          "Uniprot" ,
          "Blood.RNA...NK.cell..NX."
        )]
  
TFs <- TFs[order(TFs$Blood.RNA...NK.cell..NX., decreasing = T), ]
 
  
go <- read.delim(paste0(hpaGoUniprotPath, "QuickGO-annotations-1592213989958-20200615.tsv"), header = T)

go <- unique(go$SYMBOL)

## 780 genes common between these
# length(intersect(GPCRs$Gene, go))

## 2035 unique genes
gpr <- unique(c(as.character(GPCRs$Gene), as.character(go)))
gpr <- as.character(gpr)
gpr <- gpr[! grepl("Homo sapiens", gpr)]
gpr <- gpr[! grepl("human", gpr)]
## MGRF

uniprot <- read.csv(paste0(hpaGoUniprotPath, "7tmrlist_uniprot.csv"), 
                    stringsAsFactors = F)

uniprot <- uniprot[grepl("HUMAN", uniprot$genes), ]
uniprot <- sapply(uniprot$genes, function(x){ unlist(strsplit(x, "_"))[1]})

uniprot <- as.character(uniprot)


tars <- read.csv(paste0(hpaGoUniprotPath, "targets_and_families.csv"), 
                 stringsAsFactors = F)

tars <- tars$HGNC.symbol[tars$Type == "gpcr"]
tars <- tars[! tars == ""]
tars <- tars[!duplicated(tars)]

gpr <- unique(c(gpr, uniprot, tars))

```

# Data analysis and visualisations

## Initial Res and Exh signatures

### Define genes for CD4 analysis
Here we have put together all the signature from CD4 Trm/exhaustion/dysfunction that we collected from the literature (see Table 1 in the paper). These are 1502 genes, of which 1353 genes exist in the single cell data. These include some well known canonical Res and Exh markers.
```{r}
TcellSigsList <-
  readRDS(paste0(
    dataPath,
    "Signatures_literature/Tcell_signature_list_literature.RDS"
  ))

selGenes <-
  unique(
    unlist(TcellSigsList)
  )

```


### Cross correlation between genes
We perform cross correlation between all the 1353 genes, and then focus on the correlation between the canonical markers of CD8 Res/Exh and other genes.
```{r}

subSmartCD4 <- as.matrix(GetAssayData(cd4))

## 1353 genes
CD4Expr <- subSmartCD4[rownames(subSmartCD4) %in% selGenes, ]

corMat <- cor(t(CD4Expr), method = "spearman")


## function for flattening the cor matrix
flattenCorrMatrix <- function(cormat
) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut]
  )
}

corStat <- flattenCorrMatrix(corMat)

```

Canonical Res and Exh markers for CD8:

* Res: "CD69", "ITGAE", "ITGA1", "RGS1"
* Exh: "HAVCR2", "PDCD1", "CTLA4", "LAYN", "CXCL13", "LAG3"

Regarding the Res genes, we select genes that have correlation more than 0.3 with canonical Res markers. Note that both ZNF683 is highly expressed in a population of blood cells (annotated as CD4_GNLY in Zhang et al paper), TIGIT is also highly expressed in a population of blood cells (annotated as CD4_FOXP3 in Zhang et al paper). PRDM1 (Blimp1) is also relatively high in blood. Although we included ITGA1 and ITGAE in our canonical Res genes, and none of them had  correlation more than 0.23 with any other gene! This step results in 69 genes.

Regarding the Exh genes, we select genes that have correlation more than 0.3 with canonical Exh markers. Note that similar to CD8 results, LAG3 did not show high correlation with any of the ~1300 genes. We get 2 genes based on their correlation with HAVCR2 (TNFRSF18 and DUSP4),  only one gene based on the correlation with PDCD1 (SRGN), and no gene from CXCL13. Most genes comes from CTLA4 and LAYN. We kept LAYN in as Exh marker as it is [reported](https://www.cell.com/cell/pdf/S0092-8674(17)30596-2.pdf) in CD4 cells too. This step results in 65 genes.

As some genes are common between both Res and Exh genes obtained above, i.e. have correlations with both Res and Exh canonical markers, we remove them from both gene lists.


```{r}
trmMinGenes <- c("CD69", "ITGAE", "ITGA1", "RGS1")
exhMinGenes <- c("HAVCR2", "PDCD1", "CTLA4", "LAYN", "CXCL13", "LAG3")
```


```{r}
trmCors <- corStat %>% filter(cor > 0.3 & (row %in% trmMinGenes | column %in% trmMinGenes))
trmGenes <- unique(c(as.character(trmCors$row),
                     as.character(trmCors$column)))

texCors <- corStat %>% filter(cor > 0.3 & (row %in% exhMinGenes | column %in% exhMinGenes))
texGenes <- unique(c(as.character(texCors$row),
                     as.character(texCors$column)))

```


Initial Res genes, before removing genes common with Exh genes:
```{r}
trmGenes
```

Initial Exh genes, before removing genes common with Res genes:
```{r}
texGenes
```

Genes common between the Res and Exh gene lists:
```{r}
comGs <- intersect(texGenes, trmGenes)
comGs

trmGenes <- trmGenes[! trmGenes %in% comGs]
texGenes <- texGenes[! texGenes %in% comGs]
```


### Sparse canonical correlation
Here, we also perform sparse canonical correlation (CCA) to remove genes that cause correlations between Res and Exh gene sets, although these genes are not common between the two signatures. We visualise the cross-correlation between the Res and Exh genes before and after removing genes based on sparse CCA.

Based on the results of the CCA, we remove "BTG3"  "BIRC3" "PRDM1" from the initial Res gene list and "IKZF2" "IL2RA" "TNFRSF9" "FOXP3" from the initial Exh gene list.
```{r, fig.height=12, fig.width = 14, warning = F, message = F, cache = T,  fig.cap = "Cross-correlation between all initial Res and Exh signature genes after CCA and removal of genes causing high correlation between the two signatures"}

library(PMA)

cca_cd4_init <- CCA(
  t(subSmartCD4[rownames(subSmartCD4) %in% trmGenes, ]), 
   t(subSmartCD4[rownames(subSmartCD4) %in% texGenes, ]))

trmCorGenes <- colnames(t(subSmartCD4[rownames(subSmartCD4) %in% trmGenes, ]))[which(cca_cd4_init$u[, 1] <= -0.1)]
#  "BTG3"  "BIRC3" "PRDM1"
texCorGenes <- colnames(t(subSmartCD4[rownames(subSmartCD4) %in% texGenes, ]))[which(cca_cd4_init$v[, 1] <= -0.1)]
# "IKZF2"   "IL2RA"   "TNFRSF9" "FOXP3" 

##---- remove genes that have non-zero values in CCA:

trmGenes <- trmGenes[! trmGenes %in% trmCorGenes]
texGenes <- texGenes[! texGenes %in% texCorGenes]


CD4Expr <- subSmartCD4[unique(c(trmGenes, texGenes)), ]

corMat <- cor(t(CD4Expr), method = "spearman")


markerCols <- c("gray90", "gray20")
names(markerCols) <- c("Res", "Exh")

sampleAnnHm <-
  ComplexHeatmap::HeatmapAnnotation(Marker = as.character(c(rep(
    "Res", length(trmGenes)
  ), rep(
    "Exh", length(texGenes)
  ))),
  col = list(Marker = markerCols))

# pdf(paste0(figPath, "CrossCor_ZhangSmart_CD4_ResTexGenes_rmComGenes_rmCCA_20200928.pdf"), height = 13, width = 14)
ComplexHeatmap::Heatmap(corMat,
                        show_column_names = T, 
                        top_annotation = sampleAnnHm, 
                        col = circlize::colorRamp2(
                          c(min(corMat), 0, max(corMat)), c("yellow3", "white", "navy")), 
                        # col = viridis::plasma(100), 
                        name = "Rho")
# dev.off()
```

```{r}
saveRDS(list(trmGenes = trmGenes, texGenes = texGenes), 
        paste0(outPath, "Initial_Res_Exh_Genes_CD4_Corr_20200928.RDS"))

# trmExhGenes <- readRDS(paste0(outPath, "Initial_Res_Exh_Genes_CD4_Corr_20200928.RDS"))
# 
# trmGenes <- trmExhGenes$trmGenes
# texGenes <- trmExhGenes$texGenes
```

## Standard Seurat pipeline
```{r, warning = F, message = F, cache = T}
cd4 <- FindVariableFeatures(cd4, selection.method = "vst", nfeatures = 2000)

all.genes <- rownames(cd4)
cd4 <- ScaleData(cd4, features = all.genes)

cd4 <- RunPCA(cd4, features = VariableFeatures(object = cd4), verbose = F)

# DimHeatmap(cd4, dims = 1:30, cells = 500, balanced = TRUE)
# ElbowPlot(cd4, ndims = 40)

cd4 <- FindNeighbors(cd4, dims = 1:15)
cd4 <- FindClusters(cd4, resolution = 0.8, random.seed = 20200901)

cd4 <- RunUMAP(cd4, dims = 1:15, n.neighbors = 40, min.dist = 0.3, seed.use = 20200928)
```

### Counts vs percent mt or number of features
```{r, fig.height = 4, fig.width = 9, fig.cap = "Relationship between percent mito genes or number of detected genes and library size (from Seurat)"}
plot1 <-
  FeatureScatter(cd4,
                 feature1 = "nCount_RNA",
                 feature2 = "percent.mt",
                 group.by = "Sub_Cluster", 
                 cols = cd4Cols)
plot2 <-
  FeatureScatter(cd4,
                 feature1 = "nCount_RNA",
                 feature2 = "nFeature_RNA",
                 group.by = "Sample", 
                 cols = cols)
plot1 + plot2

```


```{r, fig.height=8, fig.width = 4, fig.cap = "Number of features, library size and percent mito genes in CD4 cells across patients"}
VlnPlot(
  cd4,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  ncol = 1,
  group.by = "Sample",
  pt.size = 0, 
  cols = cols
)
```

### Highly Variable genes
```{r, fig.cap = "Highly variable genes"}
# cd4 <- FindVariableFeatures(cd4, selection.method = "vst", nfeatures = 2000)

top10 <- head(VariableFeatures(cd4), 20)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(cd4)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```

```{r, fig.height = 6, fig.width = 6, fig.cap = "Gene loadings for PC1 and PC2"}
VizDimLoadings(cd4, dims = 1:2, reduction = "pca")
```



## Score SMART-seq2 CD4 T cells based on initial Res and Exh signatures
```{r, message = F, warning = F}
rd <- rankGenes(as.matrix(GetAssayData(cd4, slot = "data", assay = "RNA")))
     
cd4$CuratedTrmGenes <- simpleScore(rankData = rd, 
                              upSet = trmGenes, 
                              knownDirection = T, 
                              centerScore = F, 
                              dispersionFun = var)$TotalScore    
     
cd4$CuratedExhGenes <- simpleScore(rankData = rd, 
                              upSet = texGenes, 
                              knownDirection = T, 
                              centerScore = F, 
                              dispersionFun = var)$TotalScore    
## Save data after scores
# saveRDS(cd4, file = paste0(outPath, "DontPush/Zhang_SmartSeq_SeuratObj_CD4_20200928_UMAP_scores.RDS"))

```


### Scatterplot of scores (related to Figure 1)
We generate scatterplots of the Initial Res vs Exh scores. In addition to reproducing a part of Figure 1, we colour cells based on different available annotations.

```{r, fig.height = 5, fig.width = 5, fig.cap = "Scatter plot of scores comparing initial Residency and initial Exhaustion signature scores"}

cell_attrsCD4 <- cd4[[]]

lowTrmPct <- quantile(cell_attrsCD4$CuratedTrmGenes, probs = 0.6)
highTrmPct <- quantile(cell_attrsCD4$CuratedTrmGenes, probs = 0.85)

lowExhPct <- quantile(cell_attrsCD4$CuratedExhGenes, probs = 0.6)
highExhPct <- quantile(cell_attrsCD4$CuratedExhGenes, probs = 0.85)

# pdf(
#   paste0(figPath, "ScatterPlot_ZhangSmart_CD4_CuratedTRM_TEX_AllCells_lines_20200928.pdf"),
#   height = 4.3,
#   width = 4.85
# )
cell_attrsCD4 %>%
  ggplot(.,
         aes(x = CuratedTrmGenes, y = CuratedExhGenes)) +
  geom_point(alpha = 0.8, size = 1, color = "gray50") +
  geom_hline(yintercept = lowExhPct,
             col = "gray10",
             lty = 5,
             size = 1) +
  geom_hline(yintercept = highExhPct,
             col = "gray10",
             size = 1) +
  geom_vline(xintercept = lowTrmPct ,
             col = "gray10",
             size = 1) +
  geom_vline(xintercept = highTrmPct,
             col = "gray10",
              lty = 5,
             size = 1) +
  xlab("Initial residency score") +
  ylab("Initial exhaustion score") +
  currentTheme 

cell_attrsCD4 %>%
  ggplot(.,
         aes(x = CuratedTrmGenes, y = CuratedExhGenes, color = Sub_Cluster)) +
  geom_point(alpha = 0.8, size = 1) +
  scale_color_manual(values = cd4Cols) +
  geom_hline(yintercept = lowExhPct,
             col = "gray10",
             lty = 5,
             size = 1) +
  geom_hline(yintercept = highExhPct,
             col = "gray10",
             size = 1) +
  geom_vline(xintercept = lowTrmPct ,
             col = "gray10",
             size = 1) +
  geom_vline(xintercept = highTrmPct,
             col = "gray10",
              lty = 5,
             size = 1) +
  xlab("Initial residency score") +
  ylab("Initial exhaustion score") +
  currentTheme +
  guides(color = guide_legend(nrow = 4,
                              override.aes = list(size = 3)))

# cell_attrsCD8_Tissue %>%
cell_attrsCD4 %>%
  # filter(Tissue == "T") %>%
  ggplot(.,
         aes(x = CuratedTrmGenes, y = CuratedExhGenes, color = Tissue)) +
  geom_point(alpha = 0.8, size = 1) +
  scale_color_manual(values = cols) +
  geom_hline(yintercept = lowExhPct,
             col = "gray10",
             lty = 5,
             size = 1) +
  geom_hline(yintercept = highExhPct,
             col = "gray10",
             size = 1) +
  geom_vline(xintercept = lowTrmPct ,
             col = "gray10",
             size = 1) +
  geom_vline(xintercept = highTrmPct,
             col = "gray10",
              lty = 5,
             size = 1) +
  xlab("Initial residency score") +
  ylab("Initial exhaustion score") +
  currentTheme +
  guides(color = guide_legend(nrow = 4,
                              override.aes = list(size = 3)))
 
 
# cell_attrsCD8_Tissue %>%
cell_attrsCD4 %>%
  # filter(Tissue == "T") %>%
  ggplot(.,
         aes(x = CuratedTrmGenes, y = CuratedExhGenes, color = MSI)) +
  geom_point(alpha = 0.8, size = 1) +
  scale_color_manual(values = cols) +
  geom_hline(yintercept = lowExhPct,
             col = "gray10",
             lty = 5,
             size = 1) +
  geom_hline(yintercept = highExhPct,
             col = "gray10",
             size = 1) +
  geom_vline(xintercept = lowTrmPct ,
             col = "gray10",
             size = 1) +
  geom_vline(xintercept = highTrmPct,
             col = "gray10",
              lty = 5,
             size = 1) +
  xlab("Initial residency score") +
  ylab("Initial exhaustion score") +
  currentTheme +
  guides(color = guide_legend(nrow = 4,
                              override.aes = list(size = 3)))
 
# dev.off()



```


## Markers for CD4 cells
```{r}
##----------- In both Tum and normal, after removing blood cells
cell_attrsCD4_Tissue <-
  cell_attrsCD4 %>% 
  filter(Tissue != "P")

cell_attrsCD4_Tissue$TissueExhCells[cell_attrsCD4_Tissue$CuratedExhGenes >
                                      highExhPct &
                                      cell_attrsCD4_Tissue$CuratedTrmGenes <
                                      lowTrmPct] <-
  "CD4_Exh"


cell_attrsCD4_Tissue$TissueExhCells[cell_attrsCD4_Tissue$CuratedTrmGenes >
                                      highTrmPct &
                                      cell_attrsCD4_Tissue$CuratedExhGenes < lowExhPct] <-
  "CD4_Res"


cell_attrsCD4_Tissue$TissueExhCells[is.na(cell_attrsCD4_Tissue$TissueExhCells)] <- "Other"

table(cell_attrsCD4_Tissue$TissueExhCells)

# save cell names for these groups: use these in percentile comparisons

CD4_Exh_Cells <- as.character(
  cell_attrsCD4_Tissue$CellName[
    cell_attrsCD4_Tissue$TissueExhCells == "CD4_Exh"])


CD4_Res_Cells <- as.character(
  cell_attrsCD4_Tissue$CellName[
    cell_attrsCD4_Tissue$TissueExhCells == "CD4_Res"])
```

Subset the Seurat object to also remove peripheral blood cells.
```{r}
CD4Tissue <- cd4[, ! cd4$Tissue == "P" ]
# all(cell_attrsCD4_Tissue$CellName == colnames(CD4Tissue))
CD4Tissue@meta.data <- cell_attrsCD4_Tissue
CD4Tissue <- SetIdent(CD4Tissue, value = "TissueExhCells")
```

### CD4 Exh markers
```{r, cache = T}
##------- Find markers for CD4-Exh vs CD4-TRM-nonEx
markers_CD4Exh <-
  FindMarkers(
    CD4Tissue,
    ident.1 = "CD4_Exh",
    ident.2 = "CD4_Res",
    only.pos = TRUE,
    min.pct = 0.5,
    logfc.threshold = 1
  )

markers_CD4Exh$gene <- rownames(markers_CD4Exh)
markers_CD4Exh$Marker <- "CD4_Exh"

## 168 genes
DEGs_CD4Exh <- markers_CD4Exh[markers_CD4Exh$avg_logFC > 1 & markers_CD4Exh$p_val_adj < 0.05, ]

# dim(DEGs_CD4Exh)
##----------- DE using MAST
# BiocManager::install("MAST")
# library(MAST)
markers_CD4Exh_mast <-
  FindMarkers(
    CD4Tissue,
    ident.1 = "CD4_Exh",
    ident.2 = "CD4_Res",
    only.pos = TRUE,
    min.pct = 0.5,
    logfc.threshold = 1,
    test.use = "MAST"
  )

markers_CD4Exh_mast$gene <- rownames(markers_CD4Exh_mast)
markers_CD4Exh_mast$Marker <- "CD4_Exh"

# dim(markers_CD4Exh_mast)

## 214 genes
DEGs_CD4Exh_mast <- markers_CD4Exh_mast[markers_CD4Exh_mast$avg_logFC > 1 & markers_CD4Exh_mast$p_val_adj < 0.05, ]

## 168 common
# length(intersect(DEGs_CD4Exh_mast$gene, DEGs_CD4Exh$gene))

## --- Only add the one gene that came up in MAST and do not exist in 
## Wilcox test

DEGs_CD4Exh_mast <- DEGs_CD4Exh_mast[! rownames(DEGs_CD4Exh_mast) %in% rownames(DEGs_CD4Exh), ]

##--- 175 genes in total
DEGs_CD4Exh <- rbind(DEGs_CD4Exh, DEGs_CD4Exh_mast)


## Add GPCRs:
DEGs_CD4Exh$IsGPCR <- FALSE
DEGs_CD4Exh$IsGPCR[DEGs_CD4Exh$gene %in% gpr] <- TRUE
## Add TFs:
DEGs_CD4Exh$IsTF <- FALSE
DEGs_CD4Exh$IsTF[DEGs_CD4Exh$gene %in% TFs$Gene] <- TRUE

```


#### Percentile filtration
Filter those high in blood cells and also high in Res cells.
We keep genes whose 65% tile expression in Exh cells are larger than the 90%tile of their expression in blood and Res cells. This results in 83 genes.

```{r}
exprData <- as.matrix(GetAssayData(cd4, slot = "data"))

gg <-  DEGs_CD4Exh$gene
exprData_Genes <- exprData[gg, ]
cell_attrsCD4 <- cd4[[]]


Exh_Pct <-
  # rowQuantiles(exprData_Genes[, CD4_Exh_Cells], probs = 0.6)
  rowQuantiles(exprData_Genes[, CD4_Exh_Cells], probs = 0.65)

Res_Pct <-
  rowQuantiles(exprData_Genes[, CD4_Res_Cells], probs = 0.9)

Blood_Pct <-
  rowQuantiles(exprData_Genes[, cell_attrsCD4$Tissue == "P"], probs = 0.9)


pctData_cd4Exh <- data.frame(cbind(Exh_Pct,
                               Res_Pct, 
                               Blood_Pct
                               ))

pctData_cd4Exh$gene <- rownames(pctData_cd4Exh)

pctData_cd4ExhPass <- pctData_cd4Exh %>% 
  filter(Exh_Pct > Res_Pct & Exh_Pct> Blood_Pct) %>% 
  data.frame()

pctData_cd4ExhPass <-
  pctData_cd4ExhPass[order(pctData_cd4ExhPass$Exh_Pct, decreasing = T),]

DT::datatable(pctData_cd4ExhPass, filter = "top") ## 83 genes


DEGs_CD4Exh$PctPass <- FALSE
DEGs_CD4Exh$PctPass[DEGs_CD4Exh$gene %in% pctData_cd4ExhPass$gene] <- TRUE

# write.table(
#   DEGs_CD4Exh,
#   paste0(outPath, "Zhang_CD4Exh_Markers_BasedOnScores_20200928.txt"),
#   row.names = F,
#   sep = "\t"
# )
```


### CD4 Res markers
```{r}
##------- Find markers for CD4-Res vs CD4-Exh
markers_CD4Res <-
  FindMarkers(
    CD4Tissue,
    ident.1 = "CD4_Res",
    ident.2 = "CD4_Exh",
    only.pos = TRUE,
    min.pct = 0.5,
    logfc.threshold = 1
  )

markers_CD4Res$gene <- rownames(markers_CD4Res)
markers_CD4Res$Marker <- "CD4_Res"

DEGs_CD4Res <- markers_CD4Res[markers_CD4Res$avg_logFC > 1 & markers_CD4Res$p_val_adj < 0.05, ]

# 164
# dim(DEGs_CD4Res)

##------- use MAST for DE
markers_CD4Res_mast <-
  FindMarkers(
    CD4Tissue,
    ident.1 = "CD4_Res",
    ident.2 = "CD4_Exh",
    only.pos = TRUE,
    min.pct = 0.5,
    logfc.threshold = 1,
    test.use = "MAST"
  )

markers_CD4Res_mast$gene <- rownames(markers_CD4Res_mast)
markers_CD4Res_mast$Marker <- "CD4_Res"

## 157 genes
DEGs_CD4Res_mast <- markers_CD4Res_mast[markers_CD4Res_mast$avg_logFC > 1 & markers_CD4Res_mast$p_val_adj < 0.05, ]

# dim(DEGs_CD4Res_mast)

## 155 common
# length(intersect(DEGs_CD4Res_mast$gene, DEGs_CD4Res$gene))

##  10 genes came up in this method that did not come up in Wilcox!
DEGs_CD4Res_mast <- DEGs_CD4Res_mast[! rownames(DEGs_CD4Res_mast) %in% rownames(DEGs_CD4Res), ]

## 81 common genes
##--- 118 genes in total
DEGs_CD4Res <- rbind(DEGs_CD4Res, DEGs_CD4Res_mast)

## Add GPCRs:
DEGs_CD4Res$IsGPCR <- FALSE
DEGs_CD4Res$IsGPCR[DEGs_CD4Res$gene %in% gpr] <- TRUE
## Add TFs:
DEGs_CD4Res$IsTF <- FALSE
DEGs_CD4Res$IsTF[DEGs_CD4Res$gene %in% TFs$Gene] <- TRUE

```


#### Percentile filtration
Filter those high in blood and also high in Exh cells --> 31 genes

```{r}
exprData <- as.matrix(GetAssayData(cd4, slot = "data"))

gg <-  DEGs_CD4Res$gene
exprData_Genes <- exprData[gg, ]

cell_attrsCD4 <- cd4[[]]

Res_Pct <-
  rowQuantiles(exprData_Genes[, CD4_Res_Cells ], probs = 0.65)
  # rowQuantiles(exprData_Genes[, CD4_Res_Cells ], probs = 0.7)

Exh_Pct <-
  rowQuantiles(exprData_Genes[, CD4_Exh_Cells ], probs = 0.9)

Blood_Pct <-
  rowQuantiles(exprData_Genes[, cell_attrsCD4$Tissue == "P"], probs = 0.9)


pctData_cd4Trm <- data.frame(cbind(Res_Pct,
                               Exh_Pct, 
                               Blood_Pct
                               ))

pctData_cd4Trm$gene <- rownames(pctData_cd4Trm)

pctData_cd4ResPass <- pctData_cd4Trm %>% 
  filter(Res_Pct > Exh_Pct & Res_Pct > Blood_Pct) %>% 
  data.frame()

pctData_cd4ResPass <-
  pctData_cd4ResPass[order(pctData_cd4ResPass$Res_Pct, decreasing = T),]

DT::datatable(pctData_cd4ResPass, filter = "top")  ## 31 genes


DEGs_CD4Res$PctPass <- FALSE
DEGs_CD4Res$PctPass[DEGs_CD4Res$gene %in% pctData_cd4ResPass$gene] <- TRUE

# write.table(
#   DEGs_CD4Res,
#   paste0(outPath, "Zhang_CD4Res_Markers_BasedOnScores_20200928.txt"),
#   row.names = F,
#   sep = "\t"
# )

```


## Score SMART-seq2 CD4 cells using the final Res and Exh signatures
```{r}
library(singscore)
exprData <- as.matrix(GetAssayData(cd4, slot = "data"))
rd <- rankGenes(exprData)

CD4_Res <- simpleScore(
  rankData = rd,
  upSet =  pctData_cd4ResPass$gene,
  knownDirection = T,
  centerScore = F
)

CD4_Exh <- simpleScore(
  rankData = rd,
  upSet =  pctData_cd4ExhPass$gene,
  knownDirection = T,
  centerScore = F
)

cd4$CD4_Res <- CD4_Res$TotalScore
cd4$CD4_Exh <- CD4_Exh$TotalScore


## UMAP coloured by the two scores
# pdf(paste0(figPath, "UMAP_ZhangSmart_Scores_CD4_20200928.pdf"),
#     height = 4, width = 4)
# cell_attrsCD4 <- cd4[[]]
# currentRedData <- cd4@reductions[["umap"]]@cell.embeddings
# 
# cell_attrsCD4$UMAP_1 <- currentRedData[, "UMAP_1"]
# cell_attrsCD4$UMAP_2 <- currentRedData[, "UMAP_2"]
#
# for(i in c("CD4_Res", "CD4_Exh")){
#   print(
#     cell_attrsCD4 %>% 
#       arrange(!!sym(i)) %>% 
#     ggplot(., 
#            aes_string(x = "UMAP_1", 
#                    y = "UMAP_2",
#                    color = i)) +
#   geom_point(alpha = 0.8, cex = 0.5) +
#   scale_color_viridis_c() +
#   theme_dark() +
#           theme(
#         legend.spacing = unit(1.5, "mm"),
#         legend.position = "top",
#         legend.key.width = unit(0.9, "cm"), 
#         legend.title = element_text(size = 8, face = "italic"),
#         legend.text = element_text(size = 8)
#       )
#   )
# }
# dev.off()
```

### Save data after final scores
```{r}
# saveRDS(cd4, paste0(outPath, "DontPush/Zhang_SmartSeq_SeuratObj_CD4_20200928_UMAP_scores2.RDS"))
```

## PCA/UMAP using marker genes
```{r, cache = T}
gg <-  c(pctData_cd4ExhPass$gene, pctData_cd4ResPass$gene)

cd4TrmExhSig <-
  RunPCA(cd4, features = rownames(cd4)[rownames(cd4) %in% gg], verbose = F)

# ElbowPlot(cd4TrmExhSig)

# Clustering and UMAP on marker genes - including both signatures
cd4TrmExhSig <- FindNeighbors(cd4TrmExhSig, reduction = "pca", dims = 1:10)
cd4TrmExhSig <- FindClusters(cd4TrmExhSig, resolution = 0.6)

cd4TrmExhSig <- RunUMAP(cd4TrmExhSig, reduction = "pca", 
          dims = 1:10,  min.dist = 0.2, n.neighbors = 40)

# DimPlot(cd4TrmExhSig, reduction = "umap")
# DimPlot(cd4TrmExhSig, reduction = "umap", group.by = "Sub_Cluster")
# DimPlot(cd4TrmExhSig, reduction = "umap", group.by = "Tissue")
# FeaturePlot(cd4TrmExhSig, reduction = "umap", features = "CD4_Exh")
# FeaturePlot(cd4TrmExhSig, reduction = "umap", features = "CD4_Res")
```



## Trajectory analysis
```{r}
## add clusters from markers:
cd4$seurat_clusters_markers <- cd4TrmExhSig$seurat_clusters

cell_attrsCD4 <- cd4[[]]
exprData <- as.matrix(GetAssayData(cd4, slot = "data"))

rd_umap_seurat <- cd4@reductions[["umap"]]@cell.embeddings
rd_pca_seurat <- cd4@reductions[["pca"]]@cell.embeddings
```

### Generate SCE
```{r}
cd4.sce <- as.SingleCellExperiment(cd4)

reducedDims(cd4.sce) <-
  SimpleList(
    ## from Seurat
    PCA_Seurat = rd_pca_seurat,
    UMAP_Seurat = rd_umap_seurat,
    
    ## scores
    CuratedScores = as.matrix(cell_attrsCD4[, c("CuratedTrmGenes", "CuratedExhGenes")]),
    MarkerScores = as.matrix(cell_attrsCD4[, c("CD4_Res", "CD4_Exh")]),
    # FinalScores = as.matrix(cell_attrsCD4[, c("Trm", "Exh")]),
    
    ## PCs from markers
    UMAP_markers = cd4TrmExhSig@reductions[["umap"]]@cell.embeddings,
    PCA_markers = cd4TrmExhSig@reductions[["pca"]]@cell.embeddings
  )

saveRDS(cd4.sce, paste0(outPath, "DontPush/Zhang_SmartSeq_CD4_SingleCellExperiment_20200928.RDS"))
# cd4.sce <- readRDS(paste0(outPath, "DontPush/Zhang_SmartSeq_CD4_SingleCellExperiment_20200928.RDS"))
```

### Slingshot (realted to Figure 1 in the paper)
```{r, fig.cap= "Trajectory analysis in CD4 T cells using the UMAP and clusters obtained based on our Res and Exh marker genes, coloured by Res and Exh signature scores"}

source(paste0(scriptPath, "slingshot_ggplot.R"))

cd4.sce <-
  slingshot(cd4.sce, clusterLabels = 'seurat_clusters_markers', reducedDim = 'UMAP_markers')

sds <- SlingshotDataSet(cd4.sce)

textSize <- 1.4

currentThemeTrj <- theme_dark() +
  theme(
    axis.title = element_text(size = rel(textSize)),
    axis.text = element_text(angle = 0, size = rel(textSize)),
    strip.text = element_text(size = rel(textSize)),
    axis.line = element_line(colour = "black", size = 0.5),
    legend.position = 'top',
    legend.title = element_text(size = rel(textSize), face = "italic"),
    legend.text = element_text(size = rel(textSize)),
    legend.key.width = unit(2.4, 'lines'),
    ## increase the line space
    plot.title = element_text(
      face = "bold",
      size = rel(textSize),
      hjust = 0.5
    )
  )

# pdf(paste0(figPath, "Trajectory_ggplot_ZhangSmart_CD4_MarkersUMAP_ClusterMarkers_ExhScore.pdf"), height = 5, width = 8)
ggplot() +
  geom_point(size = 1) +
  gg_plot(sds, col = cd4.sce$CD4_Exh) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  scale_color_viridis_c(
    option = "viridis", 
    name = "CD4 \nexhaustion\nscore")  +
  currentThemeTrj 

ggplot() +
  geom_point(size = 1) +
  gg_plot(sds, col = cd4.sce$CD4_Res) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  scale_color_viridis_c(
    option = "viridis", 
    name = "CD4 \nresidency\nscore")  +
  currentThemeTrj 
# dev.off()

```

Check the expression of some selected genes on the trajectory
```{r, fig.cap="Trajectory analysis in CD4 T cells using the UMAP and clusters obtained based on our Res and Exh marker genes, coloured by the expression of selected genes"}

gs <-
  c("IFNG",
    "GZMK",
    "CCR7",
    "SELL",
    "GPR183",
    "GPR15",
    "PDCD1",
    "CTLA4",
    "HAVCR2",
    "TIGIT")

# pdf(
#   paste0(
#     figPath,
#     "Trajectory_ggplot_ZhangSmart_CD4_MarkersUMAP_ClusterMarkers_selGenes.pdf"
#   ),
#   height = 5,
#   width = 8
# )
for(i in gs){
  d0 <- as.matrix(assay(cd4.sce))[i,]

print(ggplot() +
  geom_point(size = 1) +
  gg_plot(sds, col = d0) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  scale_color_viridis_c(
    option = "viridis", 
    name = i)  +
  currentThemeTrj )
}
# dev.off()
```

# Filter markers based on the tumour cell data

## CCLE
"MARCH3" does not exist in CCLE data
```{r, cache = T, warning=F, message=F, fig.cap="Expression of CD4 Res and Exh genes in the CCLE  CRC data", fig.height=15, fig.width=8 }
selectedGenes <- c(pctData_cd4ResPass$gene, pctData_cd4ExhPass$gene)

dge <- readRDS(paste0(ccleLcmPath, "DGEList_CCLE_panCancer.RDS"))

library(edgeR)
  
kp <- rowSums(cpm(dge) > 1) >= 5 |
  dge$genes$Symbol %in% selectedGenes
summary(kp)

dgeFilt <- dge[kp, ]
dgeFiltNorm <- calcNormFactors(dgeFilt)

logRPKM <- rpkm(dgeFiltNorm, log = T, gene.length = dgeFiltNorm$genes$gene_length)

gg <- selectedGenes[selectedGenes %in% dgeFiltNorm$genes$Symbol]
logRPKM_CRC <- logRPKM[, dgeFiltNorm$samples$type_refined == "colorectal"]

brbg <- rev(RColorBrewer::brewer.pal(11, "PRGn"))
brbg_hcl <- colorspace::diverging_hcl(11,
  h = c(180, 50), c = 80, l = c(20, 95), power = c(0.7, 1.3))

ComplexHeatmap::Heatmap(logRPKM_CRC[gg, ! is.na(colnames(logRPKM_CRC))], 
                        col = brbg_hcl, 
                        show_column_names = F, name = "logRPKM")

```

We subset genes to those that have low expression in CCLE. These are genes whose median expression is =< 0 across samples.
```{r}
ggLogRPKM <- logRPKM_CRC[gg, ! is.na(colnames(logRPKM_CRC))]
markersPassCCLE <- rownames(ggLogRPKM)[rowMedians(ggLogRPKM) <= 0]
markersPassCCLE
```

#
## LCM (Laser captured micro-dissection)
Microarray data downloaded and processed from GSE21510.
```{r, cache = T, warning=F, message=F, fig.cap="Expression of CD8 Res and Exh genes in the LCM  CRC data", fig.height=15, fig.width=8}
ll <- readRDS(paste0(ccleLcmPath, "ColoRectal_GeneExpression_LCM_MicroArray_GSE21510_AvergedProbes_Filtered.rds"))

## 67/70 genes --> missing:  "KRT86"    "APOBEC3H" "TTC24" 
gg <- selectedGenes[selectedGenes %in% rowData(ll)$Gene.Symbol]

l <- ll[rowData(ll)$Gene.Symbol %in% gg , ll$`tissue:ch1` == "cancer, LCM" ]

lAnnot <- colData(l)

# lAnnot$Stage[grepl("1", lAnnot$`stage:ch1`)] <- "I"
# lAnnot$Stage[grepl("2", lAnnot$`stage:ch1`)] <- "II"
# lAnnot$Stage[grepl("3", lAnnot$`stage:ch1`)] <- "III"
# lAnnot$Stage[grepl("4", lAnnot$`stage:ch1`)] <- "IV"

lExpr <- assay(l)
row.names(lExpr) <- rowData(l)$Gene.Symbol

brbg <- rev(RColorBrewer::brewer.pal(11, "PRGn"))
brbg_hcl <- colorspace::diverging_hcl(11,
  h = c(180, 50), c = 80, l = c(20, 95), power = c(0.7, 1.3))

ComplexHeatmap::Heatmap(lExpr, 
                        col = brbg_hcl, 
                        show_column_names = F, name = "logExpr")

```


We also keep genes whose median expression is =< median of all genes in the data.
```{r}
th <- quantile(rowMedians(assay(ll)), probs = 0.5)
markersPassLCM <- rownames(lExpr)[rowMedians(lExpr) <= th]
markersPassLCM
#  [1] "RASGEF1B" "HAVCR2"   "CLEC7A"   "IKZF2"    "CD79B"    "LAYN"     "ACSL4"    "FNDC3B"   "PPP1R15A" "HLA-DQA1" "F5"       "IL2RA"    "IL12RB2" 
# [14] "TNFRSF9"  "ENTPD1"   "NR4A3"    "TNFRSF4"  "CCR8"     "CADM1"    "ADAM19"   "CD7"      "IL21R"    "ZBTB32"   "TNIP3"    "MZB1"     "CTLA4"   
# [27] "FOXP3"    "CD274"    "TNFRSF18" "ASB2"     "TOX2"     "TIGIT" 

# intersect(markersPassLCM, markersPassCCLE)
#  [1] "RASGEF1B" "HAVCR2"   "CLEC7A"   "IKZF2"    "CD79B"   
#  [6] "LAYN"     "HLA-DQA1" "F5"       "IL2RA"    "IL12RB2" 
# [11] "TNFRSF9"  "NR4A3"    "TNFRSF4"  "CCR8"     "CADM1"   
# [16] "ADAM19"   "CD7"      "IL21R"    "ZBTB32"   "TNIP3"   
# [21] "MZB1"     "CTLA4"    "FOXP3"    "CD274"    "TNFRSF18"
# [26] "ASB2"     "TOX2"     "TIGIT"   

# markersPassLCM[! markersPassLCM %in% markersPassCCLE]
# "ACSL4"   "FNDC3B"   "PPP1R15A" "ENTPD1" 
```



All Res and Exh genes that passed the threshold in either of the two data sets:
```{r}
passGenes <- unique(c(markersPassLCM, markersPassCCLE))
passGenes
```


Residency genes that pass the thresholds in either of the two data sets:
```{r}
ResPass <- passGenes[ passGenes %in% pctData_cd4ResPass$gene]
ResPass
```

Exhaustion genes that pass the thresholds in either of the two data sets:
```{r}
ExhPass <- passGenes[ passGenes %in% pctData_cd4ExhPass$gene]
ExhPass
```

Add these information to the signatures and export them.
```{r}

resData <- pctData_cd4ResPass

resData$CLsPass <- FALSE
resData$CLsPass[resData$gene %in% markersPassCCLE] <- TRUE

resData$LcmPass <- FALSE
resData$LcmPass[resData$gene %in% markersPassLCM] <- TRUE

resData$CancerPass <- FALSE
resData$CancerPass[resData$gene %in% ResPass] <- TRUE


exhData <- pctData_cd4ExhPass

exhData$CLsPass <- FALSE
exhData$CLsPass[exhData$gene %in% markersPassCCLE] <- TRUE

exhData$LcmPass <- FALSE
exhData$LcmPass[exhData$gene %in% markersPassLCM] <- TRUE

exhData$CancerPass <- FALSE
exhData$CancerPass[exhData$gene %in% ExhPass] <- TRUE

```


### Save merged markers
```{r}
exhData$Marker <- "Exh"
resData$Marker <- "Res"

# ResExhDataList <- list(exhData = exhData, 
#                        resData = resData)
# saveRDS(ResExhDataList, paste0(outPath, "ResExhMarkers_CD4_CLs_LCM.RDS"))

ResExhData <- rbind(resData, exhData)

write.table(
  ResExhData,
  paste0(outPath, "ResExhMarkers_CD4_CLs_LCM.txt"),
  row.names = F,
  sep = "\t"
)
```


# Session information
Here are the list of tools and packages (along with their versions) used in this document. 
```{r}
sessionInfo()
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 598px;"></div>

