compare one vs others:
https://stackoverflow.com/questions/21367259/r-cox-hazard-model-not-including-levels-of-a-factor  
# Set up and overview
In this document, we compare teh results of the survival analyses in teh TCGA to those from the Marisa et al data.
```{r}
mainDir <- getwd()
outPath <- "../output/CompareSurvival/"
figPath <- "../figure/CompareSurvival/"
dataPath <- "../data/"
scriptPath <- "../script/"

marPath <- "../output/Marisa/"
tcgaPath <- "../output/TCGA/"


ifelse(!dir.exists(file.path(mainDir, outPath)), dir.create(file.path(mainDir, outPath)), FALSE)
ifelse(!dir.exists(file.path(mainDir, figPath)), dir.create(file.path(mainDir, figPath)), FALSE)
ifelse(!dir.exists(file.path(mainDir, dataPath)), dir.create(file.path(mainDir, dataPath)), FALSE)
ifelse(!dir.exists(file.path(mainDir, scriptPath)), dir.create(file.path(mainDir, scriptPath)), FALSE)

# library(ggplot2)
library(tidyverse)
library(RColorBrewer)
library(ggbeeswarm)
library(SingleCellExperiment)
library(Seurat)
library(hgu133a.db)

options(digits = 3)

equal_breaks <- function(n = nBreak, s = scalingFactor, ...){
  function(x){
    # rescaling
    d <- s * diff(range(x)) / (1+2*s)
    round( seq(min(x)+d, max(x)-d, length=n), 2)
  }
}

nBreak = 3
scalingFactor = 0.05

textSize <- 1.3

currentTheme <- theme_bw() +
  theme(
    # panel.background = element_blank()
    axis.title = element_text(size = rel(textSize)),
    axis.text = element_text(angle = 0, size = rel(textSize)),
    # strip.background = element_rect(colour = "#f0f0f0", fill = "#f0f0f0"),
    strip.text = element_text(size = rel(textSize)),
    axis.line = element_line(colour = "black", size = 0.5),
    legend.position = 'top',
    legend.title = element_text(size = rel(textSize), face = "italic"),
    legend.text = element_text(size = rel(textSize)),
    legend.key.size = unit(1, 'lines'),
    ## increase the line space
    plot.title = element_text(
      face = "bold",
      size = rel(textSize),
      hjust = 0.5
    )
  )

cols <-  c(
  brewer.pal(8, "Dark2")[-5],
  brewer.pal(10, "Paired"),
  brewer.pal(12, "Set3"),
  brewer.pal(9, "Blues")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Oranges")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Greens")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Purples")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Reds")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Greys")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "BuGn")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "PuRd")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "BuPu")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "YlGn")[c(8, 3, 7, 4, 6, 9, 5)])


source("../script/coxph_analysis.R")
source("../script/survival_plot.R")

```


# Compare OS 
## Score pairs

```{r}
##---- without covars
os_scorePairs_mar <- read.csv(
  paste0(
    marPath,
    "Marisa_Cox_OS_ExtremeScorePairs_significant_Str_pairwise6.csv"
    # "Marisa_Cox_OS_ExtremeScorePairs_significant_Str_pairwise6_noCovars.csv"
  ),
  stringsAsFactors = F,
  check.names = F
)

os_scorePairs_tcga <-
  read.csv(
    paste0(
      tcgaPath,
      "TCGA_Cox_OS_ExtremeScorePairs_significant_Str_pairwise6.csv"
      # "TCGA_Cox_OS_ExtremeScorePairs_significant_Str_pairwise6_noCovars.csv"
    ),
    stringsAsFactors = F,
    check.names = F
  )

##--get common contrasts
comCntrasts <-
  intersect(os_scorePairs_tcga$Contrast, os_scorePairs_mar$Contrast)

com_os_scorePairs <-
  unique(os_scorePairs_tcga$VariableName[os_scorePairs_tcga$Contrast %in% comCntrasts])
com_os_scorePairs

```



### Plot survival
Generate Figure 6B, and Figure S21. The code below, also generates soem other additional surival curves that have not been included in the paper.
```{r, fig.height = 6.5, fig.width = 12}
library(gridExtra)

wide2CombLong_tcga <- readRDS(paste0(tcgaPath, "TCGA_Survival_OS_ScoreExpr_wide_Str.RDS"))
wide2CombLong_mar <- readRDS(paste0(marPath, "Marisa_Survival_OS_ScoreExpr_wide_Str.RDS"))

selScorePair_tcga <- wide2CombLong_tcga[
  wide2CombLong_tcga$Comb_Name  %in% 
    com_os_scorePairs, ]

selScorePair_mar <- wide2CombLong_mar[
  wide2CombLong_mar$Comb_Name  %in% 
    com_os_scorePairs, ]


pdf(
  # paste0(figPath, "OS_Comparison_ScorePairs_Str_Cont.pdf"),
  paste0(figPath, "OS_Comparison_ScorePairs_Str_Pairwise_ComContrasts.pdf"),
  height = 6.5,
  width = 12
)

for(j in com_os_scorePairs){
  survival_pairPlots(
  data1 = selScorePair_tcga,
  data2 = selScorePair_mar,
  strataName = j,
  strataNameColumn = "Comb_Name",
  strataColumn = "Comb_2status",
  mainTitle1 = "TCGA\n",
  mainTitle2 = "Marisa\n",
  timeCol1  = "OS.time",
  eventCol1 = "OS",
  timeCol2 = "OS.time",
  eventCol2 = "OS",
  nColLegend = 1,
  confInt = F,
  ylabel1 = "OS",
  ylabel2 = "OS",
  CoxPval = " < 0.05",
  cols = cols,
  currentTheme = currentTheme
  )
}

dev.off()

```


### Coef plot
Generate Hazard ratio plots for  significant terms in both the TCGA and Marisa data.
```{r}
# If a data frame is used, it must contain columns named term (names of predictors) and estimate (corresponding estimates of coefficients or other quantities of interest).
cox_OS_2Comb_sig_tcga <- os_scorePairs_tcga

## subset to common ones
cox_OS_2Comb_sig_tcga <- cox_OS_2Comb_sig_tcga[cox_OS_2Comb_sig_tcga$Contrast %in% comCntrasts, ]

cox_OS_2Comb_sig_tcga$Contrast <- gsub(
  "High", "H", cox_OS_2Comb_sig_tcga$Contrast
)

cox_OS_2Comb_sig_tcga$Contrast <- gsub(
  "Low", "L", cox_OS_2Comb_sig_tcga$Contrast
)


cox_OS_2Comb_sig_mar <- os_scorePairs_mar

## subset to common ones
cox_OS_2Comb_sig_mar <- cox_OS_2Comb_sig_mar[cox_OS_2Comb_sig_mar$Contrast %in% comCntrasts, ]


cox_OS_2Comb_sig_mar$Contrast <- gsub(
  "High", "H", cox_OS_2Comb_sig_mar$Contrast
)

cox_OS_2Comb_sig_mar$Contrast <- gsub(
  "Low", "L", cox_OS_2Comb_sig_mar$Contrast
)



dd_tcga <- cox_OS_2Comb_sig_tcga[,
                                 c("VariableName",
                                   # "variable",
                                   "Contrast", ## instead on variable
                                   "exp(coef)",
                                   "lower .95",
                                   "upper .95",
                                   "z",
                                   "Pr(>|z|)")]



dd_mar <- cox_OS_2Comb_sig_mar[,
                                 c("VariableName",
                                   "Contrast",
                                   # "variable",
                                   "exp(coef)",
                                   "lower .95",
                                   "upper .95",
                                   "z",
                                   "Pr(>|z|)")]

colnames(dd_tcga) <- colnames(dd_mar) <-
  # colnames(dd_tcga_Str) <- colnames(dd_mar_Str) <-
  c("VariableName",
    "term",
    "estimate",
    "conf.low",
    "conf.high",
    "z",
    "p-value")


dd_tcga <- dd_tcga[order(dd_tcga$estimate), ]
dd_mar <- dd_mar[order(dd_mar$estimate), ]


p1 <- ggstatsplot::ggcoefstats(
  x = dd_tcga, # for all significant terms in tcga
  output = "plot",
  statistic = "z",
  # point.args = list(size = 2, color = "blue"),
  vline = F,
  title = "TCGA (OS)", 
  xlab = "Hazard ratio [exp(coef)]",
  ylab = ""
) + geom_vline(xintercept = 1, lty = 2, size = 1)

p2 <- ggstatsplot::ggcoefstats(
  x = dd_mar[match(dd_tcga$term, dd_mar$term), ],
  output = "plot",  
  statistic = "z",
  vline = F,
  title = "Marisa (OS)", 
  xlab = "Hazard ratio [exp(coef)]",
  ylab = ""
) + geom_vline(xintercept = 1, lty = 2, size = 1)


library(gridExtra)

pdf(
  paste0(figPath, "OS_Comparison_ScorePairs_CoefPlot_SignifBoth_Pairwise_ComContrasts.pdf"),
  # height = 4,
  height = 4.5,
  width = 15
)

grid.arrange(p1, p2,
             ncol = 2)

dev.off()


```



### logHR - OS

Here, we convert the Hazard ratio to log Hazard ratio values to have a better visulization. Generate Supp. Figure S20 (top panel)
```{r}

ddLog_tcga <- dd_tcga
ddLog_mar <- dd_mar


ddLog_tcga$estimate <- log(ddLog_tcga$estimate)
ddLog_tcga$conf.low <- log(ddLog_tcga$conf.low)
ddLog_tcga$conf.high <- log(ddLog_tcga$conf.high)

ddLog_mar$estimate <- log(ddLog_mar$estimate)
ddLog_mar$conf.low <- log(ddLog_mar$conf.low)
ddLog_mar$conf.high <- log(ddLog_mar$conf.high)

p1 <- ggstatsplot::ggcoefstats(
  x = ddLog_tcga, # for all significant terms in tcga
  output = "plot",
  statistic = "z",
  # point.args = list(size = 2, color = "blue"),
  vline = F,
  title = "TCGA (OS)", 
  xlab = "log(Hazard ratio)",
  ylab = ""
) + geom_vline(xintercept = 0, lty = 2, size = 1)

p2 <- ggstatsplot::ggcoefstats(
  x = ddLog_mar[match(ddLog_tcga$term, ddLog_mar$term), ],
  output = "plot",  
  statistic = "z",
  vline = F,
  title = "Marisa (OS)", 
  xlab = "log(Hazard ratio)",
  ylab = ""
) + geom_vline(xintercept = 0, lty = 2, size = 1)


library(gridExtra)

pdf(
  paste0(figPath, "OS_Comparison_ScorePairs_CoefPlot_LogHR_SignifBoth_Pairwise_ComContrasts.pdf"),
  # height = 4,
  height = 4.5,
  width = 15
)

grid.arrange(p1, p2,
             ncol = 2)

dev.off()
```



# Compare PFI (TCGA) and RFS (Marisa)
## Score pairs
```{r}
rfs_scorePairs_mar <- read.csv(
  paste0(
    marPath,
    # "Marisa_Cox_RFS_ExtremeScorePairs_significant_Str_pairwise6_noCovars.csv"
    "Marisa_Cox_RFS_ExtremeScorePairs_significant_Str_pairwise6.csv"
  ),
  stringsAsFactors = F,
  check.names = F
)



pfi_scorePairs_tcga <-
  read.csv(
    paste0(
      tcgaPath,
      # "TCGA_Cox_PFI_ExtremeScorePairs_significant_Str_pairwise6_noCovars.csv"
      "TCGA_Cox_PFI_ExtremeScorePairs_significant_Str_pairwise6.csv"
    ),
    stringsAsFactors = F,
    check.names = F
  )


##---- get common contrasts
comCntrasts <-
  intersect(pfi_scorePairs_tcga$Contrast, rfs_scorePairs_mar$Contrast)

com_pfi_scorePairs <-
  unique(pfi_scorePairs_tcga$VariableName[pfi_scorePairs_tcga$Contrast %in% comCntrasts])

com_pfi_scorePairs

```

Common significant score pairs and contrasts across the two survival metric (OS and PFI/RFS) in both data sets:
```{r}
intersect(com_pfi_scorePairs, com_os_scorePairs)
# [1] "CD4_Res_Bulk--CD8NK_Exh_Bulk"
# [2] "NK_Res_Bulk--CD8_Exh_Bulk"   
# [3] "NK_Res_Bulk--NK_Exh_Bulk"    
# [4] "NK_Res_Bulk--CD8NK_Exh_Bulk" 
# [5] "CD8_Exh_Bulk--NK_Res_Bulk"   
# [6] "NK_Exh_Bulk--NK_Res_Bulk"    
# [7] "CD8NK_Exh_Bulk--CD4_Res_Bulk"
# [8] "CD8NK_Exh_Bulk--NK_Res_Bulk" 

intersect(  intersect(pfi_scorePairs_tcga$Contrast, rfs_scorePairs_mar$Contrast),   intersect(os_scorePairs_tcga$Contrast, os_scorePairs_mar$Contrast))

# [1] "Low CD4_Res_Bulk & High CD8NK_Exh_Bulk - High CD4_Res_Bulk & Low CD8NK_Exh_Bulk"
# [2] "Low NK_Res_Bulk & High CD8_Exh_Bulk - High NK_Res_Bulk & Low CD8_Exh_Bulk"      
# [3] "Low NK_Res_Bulk & High NK_Exh_Bulk - High NK_Res_Bulk & Low NK_Exh_Bulk"        
# [4] "Low NK_Res_Bulk & High CD8NK_Exh_Bulk - High NK_Res_Bulk & Low CD8NK_Exh_Bulk"  
# [5] "Low CD8_Exh_Bulk & High NK_Res_Bulk - High CD8_Exh_Bulk & Low NK_Res_Bulk"      
# [6] "Low NK_Exh_Bulk & High NK_Res_Bulk - High NK_Exh_Bulk & Low NK_Res_Bulk"        
# [7] "Low CD8NK_Exh_Bulk & High CD4_Res_Bulk - High CD8NK_Exh_Bulk & Low CD4_Res_Bulk"
# [8] "Low CD8NK_Exh_Bulk & High NK_Res_Bulk - High CD8NK_Exh_Bulk & Low NK_Res_Bulk" 
```

### Plot survival
Generate Supp. Figure S22 and S23, along with survival curves for other significant score pairs.
```{r}
library(gridExtra)
wide2CombLong_tcga <- readRDS(paste0(tcgaPath, "TCGA_Survival_PFI_ScoreExpr_wide_Str.RDS"))
wide2CombLong_tcga$PFI.time <- wide2CombLong_tcga$PFI.time/30

wide2CombLong_mar <- readRDS(paste0(marPath, "Marisa_Survival_RFS_ScoreExpr_wide_Str.RDS"))

selScorePair_tcga <- wide2CombLong_tcga[
  wide2CombLong_tcga$Comb_Name  %in% 
    com_pfi_scorePairs, ]

selScorePair_mar <- wide2CombLong_mar[
  wide2CombLong_mar$Comb_Name  %in% 
    com_pfi_scorePairs, ]



pdf(
  paste0(figPath, "PFI_RFS_Comparison_ScorePairs_signifBoth_Str_Pairwise_ComContrasts.pdf"),
  height = 6.5,
  width = 12
)

for(j in com_pfi_scorePairs){
  survival_pairPlots(
  data1 = selScorePair_tcga,
  data2 = selScorePair_mar,
  strataName = j,
  strataNameColumn = "Comb_Name",
  strataColumn = "Comb_2status",
  mainTitle1 = "TCGA\n",
  mainTitle2 = "Marisa\n",
  timeCol1  = "PFI.time",
  eventCol1 = "PFI",
  timeCol2 = "RFS.time",
  eventCol2 = "RFS",
  nColLegend = 1,
  confInt = F,
  ylabel1 = "PFI",
  ylabel2 = "RFS",
  CoxPval = " < 0.05",
  cols = cols,
  currentTheme = currentTheme
  )
}
  
dev.off()

```

### Coef plot
Hazard ratio plot using PFI and RFS data.
```{r}
cox_pfi_2Comb_sig_tcga <- pfi_scorePairs_tcga

## subset to common ones
cox_pfi_2Comb_sig_tcga <- cox_pfi_2Comb_sig_tcga[cox_pfi_2Comb_sig_tcga$Contrast %in% comCntrasts, ]

cox_pfi_2Comb_sig_tcga$Contrast <- gsub(
  "High", "H", cox_pfi_2Comb_sig_tcga$Contrast
)

cox_pfi_2Comb_sig_tcga$Contrast <- gsub(
  "Low", "L", cox_pfi_2Comb_sig_tcga$Contrast
)

cox_rfs_2Comb_sig_mar <- rfs_scorePairs_mar

## subset to common ones
cox_rfs_2Comb_sig_mar <- cox_rfs_2Comb_sig_mar[cox_rfs_2Comb_sig_mar$Contrast %in% comCntrasts, ]


cox_rfs_2Comb_sig_mar$Contrast <- gsub(
  "High", "H", cox_rfs_2Comb_sig_mar$Contrast
)

cox_rfs_2Comb_sig_mar$Contrast <- gsub(
  "Low", "L", cox_rfs_2Comb_sig_mar$Contrast
)


dd_tcga <- cox_pfi_2Comb_sig_tcga[,
                                 c("VariableName",
                                   "Contrast",
                                   "exp(coef)",
                                   "lower .95",
                                   "upper .95",
                                   "z",
                                   "Pr(>|z|)")]



dd_mar <- cox_rfs_2Comb_sig_mar[,
                                 c("VariableName",
                                   "Contrast",
                                   "exp(coef)",
                                   "lower .95",
                                   "upper .95",
                                   "z",
                                   "Pr(>|z|)")]

colnames(dd_tcga) <- colnames(dd_mar) <-
  c("VariableName",
    "term",
    "estimate",
    "conf.low",
    "conf.high",
    "z",
    "p-value")



dd_tcga <- dd_tcga[order(dd_tcga$estimate), ]
dd_mar <- dd_mar[order(dd_mar$estimate), ]

p1 <- ggstatsplot::ggcoefstats(
  x = dd_tcga, # for all significant terms in tcga
  output = "plot",
  statistic = "z",
  # point.args = list(size = 2, color = "blue"),
  vline = F,
  title = "TCGA (PFI)", 
  xlab = "Hazard ratio [exp(coef)]",
  ylab = ""
) + geom_vline(xintercept = 1, lty = 2, size = 1)

p2 <- ggstatsplot::ggcoefstats(
  x = dd_mar[match(dd_tcga$term, dd_mar$term), ],
  output = "plot",  
  statistic = "z",
  vline = F,
  title = "Marisa (RFS)", 
  xlab = "Hazard ratio [exp(coef)]",
  ylab = ""
) + geom_vline(xintercept = 1, lty = 2, size = 1)


pdf(
  paste0(figPath, "PFI_RFS_Comparison_ScorePairs_CoefPlot_SignifBoth_Pairwise_ComContrasts.pdf"),
  height = 6,
  # height = 4,
  width = 15
)

grid.arrange(p1, p2,
             ncol = 2)

dev.off()

```

### logHR - PFI
Generate Supp. Figure S20 (bottom panel) containing log hazard ratios. 
```{r}
ddLog_tcga <- dd_tcga
ddLog_mar <- dd_mar


ddLog_tcga$estimate <- log(ddLog_tcga$estimate)
ddLog_tcga$conf.low <- log(ddLog_tcga$conf.low)
ddLog_tcga$conf.high <- log(ddLog_tcga$conf.high)

ddLog_mar$estimate <- log(ddLog_mar$estimate)
ddLog_mar$conf.low <- log(ddLog_mar$conf.low)
ddLog_mar$conf.high <- log(ddLog_mar$conf.high)

p1 <- ggstatsplot::ggcoefstats(
  x = ddLog_tcga, # for all significant terms in tcga
  output = "plot",
  statistic = "z",
  # point.args = list(size = 2, color = "blue"),
  vline = F,
  title = "TCGA (PFI)", 
  xlab = "log(Hazard ratio)",
  ylab = ""
) + geom_vline(xintercept = 0, lty = 2, size = 1)

p2 <- ggstatsplot::ggcoefstats(
  x = ddLog_mar[match(ddLog_tcga$term, ddLog_mar$term), ],
  output = "plot",  
  statistic = "z",
  vline = F,
  title = "Marisa (RFS)", 
  xlab = "log(Hazard ratio)",
  ylab = ""
) + geom_vline(xintercept = 0, lty = 2, size = 1)


library(gridExtra)

pdf(
  paste0(figPath, "PFI_RFS_Comparison_ScorePairs_CoefPlot_LogHR_SignifBoth_Pairwise_ComContrasts.pdf"),
  # height = 4,
  height = 4.5,
  width = 15
)

grid.arrange(p1, p2,
             ncol = 2)

dev.off()
```

# Session information
Here are the list of tools and packages (along with their versions) used in this document. 
```{r}
sessionInfo()
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 598px;"></div>
    
      


