---
title: 'Survival and mutation analyses of the Marisa et al COAD data based on the Res and Exh scores'
author: 
  - name: 'Momeneh (Sepideh) Foroutan and Ramyar Molania'
    affiliation: 'Huntington Cancer Immunotherapy Lab'
    url: https://www.monash.edu/discovery-institute/huntington-lab
date: '12-06-2020'
output:
  rmdformats::readthedown:
    fig_width: 12
    fig_height: 6
    gallery: TRUE
    highlight: tango
    lightbox: TRUE
    self_contained: TRUE
    thumbnails: FALSE
    number_sections: TRUE	
    toc_depth: 4
    use_bookdown: TRUE
    code_folding: hide
  html_document2:
    df_print: paged
params:
  update_date: !r paste("Last updated on:", Sys.Date() )
editor_options: 
  chunk_output_type: inline
---
`r params$update_date`

<style>
body {
text-align: justify}
</style>

# Set up and overview
In this document, we read and process the COAD microarray data (GSE39582) from [Marisa et al](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3660251/), we perform CMS clustering, and visualise the associations between Res_bulk and Exh_bulk scores and different sample annotations. We then perform survival analysis to examine the association of the scores with survival outcomes.
```{r}
mainDir <- getwd()
outPath <- "../output/Marisa/"
figPath <- "../figure/Marisa/"
dataPath <- "../data/"
scriptPath <- "../script/"

marisaPath <- "../data/Marisa/"

ifelse(!dir.exists(file.path(mainDir, outPath)), dir.create(file.path(mainDir, outPath)), FALSE)
ifelse(!dir.exists(file.path(mainDir, figPath)), dir.create(file.path(mainDir, figPath)), FALSE)
ifelse(!dir.exists(file.path(mainDir, dataPath)), dir.create(file.path(mainDir, dataPath)), FALSE)
ifelse(!dir.exists(file.path(mainDir, scriptPath)), dir.create(file.path(mainDir, scriptPath)), FALSE)

# library(ggplot2)
library(tidyverse)
library(RColorBrewer)
library(ggbeeswarm)
library(SingleCellExperiment)
library(Seurat)
library(hgu133a.db)

options(digits = 3)

equal_breaks <- function(n = nBreak, s = scalingFactor, ...){
  function(x){
    # rescaling
    d <- s * diff(range(x)) / (1+2*s)
    round( seq(min(x)+d, max(x)-d, length=n), 2)
  }
}

nBreak = 3
scalingFactor = 0.05

textSize <- 1.2

currentTheme <- theme_bw() +
  theme(
    # panel.background = element_blank()
    axis.title = element_text(size = rel(textSize)),
    axis.text = element_text(angle = 0, size = rel(textSize)),
    # strip.background = element_rect(colour = "#f0f0f0", fill = "#f0f0f0"),
    strip.text = element_text(size = rel(textSize)),
    axis.line = element_line(colour = "black", size = 0.5),
    legend.position = 'top',
    legend.title = element_text(size = rel(textSize), face = "italic"),
    legend.text = element_text(size = rel(textSize)),
    legend.key.size = unit(1, 'lines'),
    ## increase the line space
    plot.title = element_text(
      face = "plain",
      size = rel(textSize),
      hjust = 0.5
    )
  )

cols <-  c(
  brewer.pal(8, "Dark2")[-5],
  brewer.pal(10, "Paired"),
  brewer.pal(12, "Set3"),
  brewer.pal(9, "Blues")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Oranges")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Greens")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Purples")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Reds")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Greys")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "BuGn")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "PuRd")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "BuPu")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "YlGn")[c(8, 3, 7, 4, 6, 9, 5)])


source(paste0(scriptPath, "visReduction_function.R"))
source("../script/coxph_analysis.R")
source("../script/survival_plot.R")
```

## Read signatures
We read Res and Exh signatures that we obtained and focus on the genes that are not highly expressed in cancer cells (those that passed the bulk cancer threshold). Additionally, we defined two other types of signatures for eacj of the Exh and Res signatures: All_Res[Exh]_Bulk and CD8NK_Res[Exh]_Bulk by merging all Res[Exh]_Bulk signature genes and by merging only the CD8_Res[Exh]_Bulk and NK_Res[Exh]_Bulk. 
We also read several cancer related signatures as well as a few other immune related signatures collected from the literature as well as MSigDB. Here are the list of the signatures that we used in this document.
```{r}
source(paste0(scriptPath, "ReadSignaturesTum_paper.R"))
names(AllSigsCollection)
```

# Marisa data
We use the GEOquery package to download these data on the 18th of March 2019 (54674 features, 585 samples). Then, the SummarizedExperiment (SE) object was generated using the SummarizedExperiment Bioconductor package. Here, we comment all the code related to the download and generation of the SE and the annotation file, and directly read the RDS file containing the processed expression data and the annotation file.

```{r}
# library(GEOquery)
# library(Biobase)
# ## download GSE39582_series_matrix.txt.gz which has 164.2 MB 
# gse <- getGEO("GSE39582", GSEMatrix = TRUE)
# show(gse)
# 
## Extract sample and gene annotations, as well as expression from the GSE. 
## We also deal with the missing row in the data.
# ##----- sample annot
# gseAnnot <- pData(phenoData(gse[[1]]))
# ## row names in annotation data are wrongly one of the expression rows
# row.names(gseAnnot) <- gseAnnot$geo_accession
# 
# ## Clean the annotation file:
# # colnames(gseAnnot)[colnames(gseAnnot) == "characteristics_ch1.11"] <- "rfs.event"
# 
# 
# ##----- Gene annot
# gseGeneAnnot <- fData(gse[[1]])
# gseGeneAnnot <-
#   gseGeneAnnot[, c(
#     "ID",
#     "Gene Symbol",
#     "ENTREZ_GENE_ID",
#     "Gene Title",
#     "GB_ACC",
#     "Annotation Date",
#     "Sequence Source"
#   )]
# 
# ##------ Expression matrix
# gseExpr <- as.data.frame(exprs(gse[[1]]))
# 
# 
# ## Colnames in expression data are wrongly one of the expression rows,
# ## same as in the sample annotation data
# all(row.names(pData(phenoData(gse[[1]]))) == colnames(gseExpr))
# 
# ## Add the missing probe expression to the expression data
# missingProbe <- as.data.frame(matrix(as.numeric(colnames(gseExpr)), nrow = 1))
# row.names(missingProbe) <- "1007_s_at"
# colnames(missingProbe)  <- row.names(gseAnnot)
#  
# colnames(gseExpr) <- row.names(gseAnnot)
# gseExpr <- rbind(missingProbe, gseExpr)
# 
# ## Add the missing gene annotation to the gene annotation data
# missingGeneAnnot <- data.frame(
#   "ID" = "1007_s_at",
#   "Gene Symbol" = "DDR1",
#   "ENTREZ_GENE_ID" = "780",
#   "Gene Title" = "Discoidin Domain Receptor Tyrosine Kinase 1",
#   "GB_ACC" = "NA", 
#   "Annotation Date" = "Oct 6, 2014",
#   "Sequence Source" = "NA"
#   , check.names = F)
# row.names(missingGeneAnnot) <- "1007_s_at"
# gseGeneAnnot <- rbind(missingGeneAnnot, gseGeneAnnot)
# 
# library(SummarizedExperiment)
# gse_se <- SummarizedExperiment(
#   assays = list(rma = as.matrix(gseExpr)), 
#   colData = DataFrame(gseAnnot), 
#   rowData = DataFrame(gseGeneAnnot))
# 
# saveRDS(gse_se, paste0(marisaPath, "GSE39582_SummExpr.RDS"))


# library(hgu133plus2.db)
# library(SummarizedExperiment)
# mr <- readRDS(paste0(marisaPath, "GSE39582/GSE39582_SummExpr.RDS"))
# assay(mr)[1:4,1:5]
# 
# mr <- mr[! grepl("///", rowData(mr)$Gene.Symbol), ]
# 
# mrExpr <- data.frame(as.matrix(assay(mr)))
# mrAnn <- rowData(mr)
# mrAnn$Gene.Symbol <- as.character(mrAnn$Gene.Symbol)
# mrExpr$Symbol <- mrAnn$Gene.Symbol
# 
# eMr <- mrExpr %>% 
#   group_by(Symbol) %>% 
#   summarise_all(median)
# 
# eMar <- as.matrix(eMr[-1, 2:ncol(eMr)])
# rownames(eMar) <- eMr$Symbol[-1]
# 
# ###---- annotaion:
# annMar <- colData(mr)
# annMar$source_name_ch1 <- as.character(annMar$source_name_ch1)
# annMar$Sample_Type[grepl("non tumoral", annMar$source_name_ch1)] <- "Normal"
# annMar$Sample_Type[grepl("primary colorectal Adenocarcinoma", annMar$source_name_ch1)] <- "Cancer"
# 
# MarisaDataList <- list(eMar = eMar, annMar = annMar)

# saveRDS(MarisaDataList, paste0(marisaPath, "DontPush/Marisa_Expr_Annot.RDS"))

MarisaDataList <- readRDS(paste0(marisaPath, "DontPush/Marisa_Expr_Annot.RDS"))
eMar <- MarisaDataList$eMar
annMar <- MarisaDataList$annMar

all(rownames(annMar) == colnames(eMar))


colnames(annMar)[colnames(annMar) == "braf.mutation.ch1"] <- "BRAF"
colnames(annMar)[colnames(annMar) == "kras.mutation.ch1"] <- "KRAS"
colnames(annMar)[colnames(annMar) == "tp53.mutation.ch1"] <- "TP53"

colnames(annMar)[colnames(annMar) == "tnm.stage.ch1"] <- "Stage"
annMar$Stage[annMar$Sample_Type == "Normal"] <- "N"
annMar$Stage[annMar$Stage == 0] <- NA

annMar$Stage[annMar$Stage == 1] <- "I"
annMar$Stage[annMar$Stage == 2] <- "II"
annMar$Stage[annMar$Stage == 3] <- "III"
annMar$Stage[annMar$Stage == 4] <- "IV"


##----- filter for Cancer samples only:
eMarFilt <- eMar[, annMar$Sample_Type == "Cancer"]
annMarFilt <- annMar[ annMar$Sample_Type == "Cancer", ]

### annMarFilt <- annMarFilt[annMarFilt$tnm.stage.ch1 != "0", ]

annMarFilt$os.delay.ch1 <- as.numeric(annMarFilt$os.delay.ch1)

colnames(annMarFilt)[colnames(annMarFilt) == "os.delay.ch1"] <- "OS.time"
colnames(annMarFilt)[colnames(annMarFilt) == "os.event.ch1"] <- "OS"
colnames(annMarFilt)[colnames(annMarFilt) == "rfs.delay.ch1"] <- "RFS.time"
colnames(annMarFilt)[colnames(annMarFilt) == "rfs.event.ch1"] <- "RFS"

colnames(annMarFilt)[colnames(annMarFilt) == "mmr.status.ch1"] <- "MMR"
colnames(annMarFilt)[colnames(annMarFilt) == "cin.status.ch1"] <- "CIN"
colnames(annMarFilt)[colnames(annMarFilt) == "cimp.status.ch1"] <- "CIMP"
colnames(annMarFilt)[colnames(annMarFilt) == "age.at.diagnosis..year..ch1"] <- "Age"


annMarFilt$Age <- as.numeric(annMarFilt$Age)
annMarFilt$OS <- as.numeric(annMarFilt$OS)
annMarFilt$OS.time <- as.numeric(annMarFilt$OS.time)
annMarFilt$RFS <- as.numeric(annMarFilt$RFS)
annMarFilt$RFS.time <- as.numeric(annMarFilt$RFS.time)

annMarFilt$Age_Status[annMarFilt$Age > median(annMarFilt$Age, na.rm = T)] <- "Old"
annMarFilt$Age_Status[annMarFilt$Age <= median(annMarFilt$Age, na.rm = T)] <- "Young"

```

## CMS
Perform CMS clustering only on Cancer samples, and add this to the annotation data.
```{r}
# eMarFilt <- eMar[, annMar$Sample_Type == "Cancer"]
# annMarFilt <- annMar[ annMar$Sample_Type == "Cancer", ]

set.seed(20201011)

cmsMar <-
  CMScaller::CMScaller(
    emat = eMarFilt,
    RNAseq = F,
    FDR = 0.05,
    rowNames = "symbol"
  )
# CMS1 CMS2 CMS3 CMS4 <NA> 
#   84  175   83  154   70 

annMarFilt$CMS <- as.character(cmsMar$prediction)

## add CMS to the annotation data including normal samples
annMar <-
  merge(
    annMar,
    cmsMar[, "prediction", drop = F],
    by.x = "geo_accession",
    by.y = "row.names",
    all.x = T
  )

colnames(annMar)[colnames(annMar) == "prediction"] <- "CMS"
annMar$CMS <- as.character(annMar$CMS)

annMar$CMS[annMar$Sample_Type == "Normal"] <- "N"

row.names(annMar) <- annMar$geo_accession
annMar <- annMar[colnames(eMar), ]

# all(rownames(annMar) == colnames(eMar))

```

## Score samples
When filtering and processing the data, we made sure that we are keeping genes (mostly immune/cancer related genes) that we also retained during the analysis and processing the TCGA COAD data. We have exported these genes as an RDS file and use that here. 
```{r}
# gg <- unique(
#   c(
#     allMarkers$gene[allMarkers$CancerPass],
#     "TGFB1", "TGFB2", "TGFB3", "TGFBR1", "TGFBR2",
#     "IL15", "IL12A", "IL12B", "IL18", "IL2",
#     "IFNG", "HIF1A", "FAS", "FASLG",
#     # Hypoxia_Buffa,
#     intersect(checkSigID$Symbol[checkSigID$geneset == "HALLMARK_HYPOXIA"],
#               Hypoxia_Buffa),
#     intersect(checkSigID$Symbol[checkSigID$geneset == "HALLMARK_HYPOXIA"],
#               sigsTum$Hypoxia),
#     CanEvasion 
#   )
# )

kpGenes <- readRDS(paste0(marisaPath, "kpGenes.RDS"))

kp1 <- rowSums(eMar > 4) >= 0.5 * ncol(eMar)
kp2 <- rownames(eMar)  %in% unique(
  c(
    kpGenes
    # gg
  )
)


eMar <- eMar[ kp1 | kp2, ]


rdMar <- rankGenes(eMar)

marScoresDisp <- multiScore(rdMar, 
                         upSetColc = AllSigsCollection, 
                         knownDirection = T, 
                         centerScore = F)

marScores <-
  data.frame(t(marScoresDisp$Scores),
             check.rows = F,
             check.names = F)

## Add TGFb-EMT bidirectional scores too:
marScores$TGFbEMT_Dir <- simpleScore(
  rdMar,
  upSet = sigsTum$TGFb_EMT$tgfb_Up,
  downSet = sigsTum$TGFb_EMT$tgfb_Dn,
  knownDirection = T,
  centerScore = F
)$TotalScore

scoreNames <- c(colnames(marScores))
scoreNames <- scoreNames[order(scoreNames)]


##----- Add these to annotation data:
annMar <-
  data.frame(annMar ,
             marScores,
             check.names = F,
             check.rows = F)


annMarFilt <- data.frame(annMarFilt,
                          annMar[row.names(annMarFilt), scoreNames], 
                          check.names = F)


```


## Score for Cibersort 
We also score for general immune and stroma signatures
```{r}
source(paste0(scriptPath, "ReadSignatures_Cibersort_TCGA.R"))

marScoresDisp_imm <- multiScore(rdMar, 
                         upSetColc = ImmSigsCollection, 
                         knownDirection = T, 
                         centerScore = F)

marScores_imm <-
  data.frame(t(marScoresDisp_imm$Scores),
             check.rows = F,
             check.names = F)



##----- Add these to annotation data:
annMar_imm <-
  data.frame(annMar ,
             marScores_imm,
             check.names = F,
             check.rows = F)

annMarFilt_imm <- data.frame(annMarFilt,
                          annMar_imm[row.names(annMarFilt), names(ImmSigsCollection)], 
                          check.names = F)
```


# Save data after scores
```{r}

# saveRDS(
#   list(
#     eMar = eMar,
#     annMar = annMar,
#     annMarFilt_imm = annMarFilt_imm
#   ),
#   paste0(outPath, "Marisa_Annot.RDS")
# )

mm <- readRDS(paste0(outPath, "Marisa_Annot.RDS"))
annMarFilt_imm <- mm$annMarFilt_imm
```


### Define Cancer and Immune signatures
```{r}
canSigs <- scoreNames[! grepl("Res_", scoreNames) &
                        ! grepl("Exh_", scoreNames)]

imSigs <- scoreNames[grepl("Res_", scoreNames) |
                        grepl("Exh_", scoreNames)]
 
```

```{r}

cibAnnot_bulk <- annMarFilt_imm[,
          (grepl("Bulk", colnames(annMarFilt_imm)) &       
         ! grepl("All_|CD8NK_", colnames(annMarFilt_imm))) |
           colnames(annMarFilt_imm) %in% cibeNames ]
           

cibAnnot_bulkLong <-
  pivot_longer(
    cibAnnot_bulk,
    cols = 7:30,
    names_to = "CIBERSORT_signatures",
    values_to = "CIBERSORT_scores"
  ) %>% 
  data.frame()

cibAnnot_bulkLong2 <- cibAnnot_bulkLong %>%
  pivot_longer(cols = 1:6,
               names_to = "ResExh_signatures",
               values_to = "ResExh_scores") %>% 
  data.frame()


pdf(
  paste0(figPath, "Scatterplot_Marisa_COBERSORT_ImmuneStroma_ResExh_Scores_Str_scaleFree.pdf"),
  # paste0(figPath, "Scatterplot_TCGA_COBERSORT_ResExh_Scores_Str.pdf"),
  height = 7,
  # width = 9
  width = 12
)
sigsResExh <- unique(cibAnnot_bulkLong2$ResExh_signatures)[
  order(unique(cibAnnot_bulkLong2$ResExh_signatures))
]
for(i in sigsResExh) {
  print(
    cibAnnot_bulkLong2 %>%
    filter(cibAnnot_bulkLong2$ResExh_signatures == i) %>%
    ggplot(., aes(ResExh_scores, CIBERSORT_scores)) +
    geom_point(color = "gray40", alpha = 0.5, size = 0.5) +
    # facet_wrap(~ CIBERSORT_signatures, ncol = 5) +
    facet_wrap(~ CIBERSORT_signatures, ncol = 6, scales = "free") +
    ggpubr::stat_cor(method = "spearman") +
    ggtitle(i) +
    theme_bw()
  )
}
dev.off()
```


We also generate a heatmap of the correlation values between Res/Exh scores and several immune signature scores from CIBERSORT. This generates Figure S14 - right panel.
```{r}

corData <- data.frame(matrix(nrow = 24, ncol = 0))
rownames(corData) <- cibeNames

imSigs_sub <- c(
  "CD4_Exh_Bulk",
  "CD4_Res_Bulk",
  "CD8_Exh_Bulk",
  "CD8_Res_Bulk",
  "NK_Exh_Bulk",
  "NK_Res_Bulk"
) 

for (im in imSigs_sub) {
      corValues <- vector()
      
  for (se in cibeNames) {
    
    corVal <- round(cor.test(annMarFilt_imm[, im],
                             annMarFilt_imm[, se])[[4]],
                    2)
    names(corVal) <- se
    corValues <- c(corValues, corVal)
    dd <- data.frame(corValues)
  }
  
  corData <- cbind(corData, dd)
}

# colnames(corData) <- selCanSigs
colnames(corData) <- imSigs_sub

library(circlize)
library(grid)

png(
  paste0(figPath, "Correlation_ResExh_CIBERSORT_Marisa_noClust.png"),
  height = 8.5,
  width = 6,
  res = 400,
  units = "in"
)
ComplexHeatmap::Heatmap(
  corData,
  # cluster_columns = T,
  cluster_columns = F,
  cluster_rows = F,
  col = viridis::viridis(100),
  column_title = "Correlation of Res and Exh scores\nwith CIBERSORT signature scores\nin Marisa COAD data",
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", corData[i, j]), x, y, gp = gpar(fontsize = 10))
  }, show_heatmap_legend = T, name = "Pearson's\ncorrelation\ncoefficient\n"
)
dev.off()

```


# Scatterplot of score pairs
Generate scatterplots of scores vs each other. Uncomment the commented sections to change the colour the data points based on CMS or MSI. 
The below code generates many scatterplots comparing each signature with all other signatures; some of these regenerates a number of figures presented in the paper:

```{r}
colnames(annMar)[colnames(annMar) == "mmr.status.ch1"] <- "MMR"  
annMar$MMR <- as.character(annMar$MMR)
annMar$MMR[annMar$MMR  %in% "N/A"] <- NA

cmsCol <- c(
  brewer.pal(11, "BrBG")[3],
  brewer.pal(11, "PRGn")[2],
  brewer.pal(9, "Blues")[5],
  brewer.pal(11, "BrBG")[10]
)


pdf(
  paste0(
    figPath,
    "ScatterPlots_Marisa_ExhResScorePairs_colCMS_rmNormal_Lines.pdf"
    # "ScatterPlots_Marisa_ExhResScorePairs_colMMR_rmNormal_Lines.pdf"
  ),
  height = 4.7,
  width = 4.3
)

      palette <- cmsCol
      # palette <- rev(c(brewer.pal(9, "Purples") [c( 6, 9)]))
      
      pal <- list(values=palette, na.value="gray80")

for(im in imSigs) {
  Others <- imSigs[imSigs != im]

  for (ot in Others) {

    d <- annMar %>%
      filter(! annMar$Sample_Type  %in%  "Normal") %>%
      mutate(medIm = median(!!sym(im)),
             medOt = median(!!sym(ot)))

    corVal <- round(cor.test(d[, im], d[, ot])[[4]], 2)

    print(
      d %>%
        ggplot(.,
               # aes_string(im, ot, color = "MMR")) +
               aes_string(im, ot, color = "CMS")) +
        geom_point(alpha = 0.8) +
        invoke(scale_colour_manual, pal) +
        geom_vline(xintercept = d$medIm) +
        geom_hline(yintercept = d$medOt) +
        scale_x_continuous(
                       breaks = equal_breaks(n = nBreak, s = scalingFactor),
                       expand = c(scalingFactor, 0)
                     ) +
        scale_y_continuous(
                       breaks = equal_breaks(n = nBreak, s = scalingFactor),
                       expand = c(scalingFactor, 0)
                     ) +
        ggtitle(paste0("Marisa (cor = ", corVal, ")")) +
        currentTheme +
        guides(color = guide_legend(nrow=2))
    )
  }
}

dev.off()






pdf(
  paste0(
    figPath,
    "ScatterPlots_Marisa_ImmunAllCancerScorePairs_colCMS_rmNormal_Lines.pdf"
  ),
  height = 4.7,
  width = 4.3
)

for(im in imSigs[!grepl("All", imSigs)]) {

  for (ca in canSigs) {
    
    d <- annMar %>%
      filter(! annMar$Sample_Type  %in%  "Normal") %>%
      mutate(medIm = median(!!sym(im)),
             medOt = median(!!sym(ca)))

    corVal <- round(
      cor.test(d[, im], d[, ca])[[4]], 2)


    palette <- cmsCol
    # palette <- rev(c(brewer.pal(9, "Purples") [c(4, 6, 9)]))

    pal <- list(values=palette, na.value="gray80")

    print(
      d %>%
        ggplot(.,
               aes_string(im, ca, color = "CMS")) +
               # aes_string(im, ca, color = "MSI")) +
        geom_point(alpha = 0.8) +
        invoke(scale_colour_manual, pal) +
        geom_vline(xintercept = d$medIm) +
        geom_hline(yintercept = d$medOt) +
        scale_x_continuous(
                       breaks = equal_breaks(n = nBreak, s = scalingFactor),
                       expand = c(scalingFactor, 0)
                     ) +
        scale_y_continuous(
                       breaks = equal_breaks(n = nBreak, s = scalingFactor),
                       expand = c(scalingFactor, 0)
                     ) +
        ggtitle(paste0("Marisa (cor = ", corVal, ")")) +
        currentTheme +
        guides(color = guide_legend(nrow=2))
    )
  }
}

dev.off()


```


We also generate a heatmap of the correlation values between Res/Exh scores and several selected signature scores. 
```{r}

corData <- data.frame(matrix(nrow = 8, ncol = 0))
rownames(corData) <- imSigs[!grepl("All", imSigs)]

# selCanSigs <-
#   c(
#     "TGFb_Signaling",
#     "TGFbEMT",
#     "HPX_EGF_EMT",
#     "OXPHOS",
#     "DDR",
#     "DNA_Repair"
#   )

# for(se in selCanSigs) {
for(se in canSigs) {
  corValues <- vector()
  for (im in imSigs[!grepl("All", imSigs)]) {
    d <- annMar %>%
      filter(! annMar$Sample_Type  %in%  "Normal")
    
    corVal <- round(cor.test(d[, im], d[, se])[[4]], 2)
    names(corVal) <- im
    corValues <- c(corValues, corVal)
    dd <- data.frame(corValues)
  }
  
  corData <- cbind(corData, dd)
}


# colnames(corData) <- selCanSigs
colnames(corData) <- canSigs

library(circlize)
library(grid)


png(
  paste0(figPath, "Correlation_ResExh_AllCanSigs_Marisa.png"),
  # paste0(figPath, "Correlation_ResExh_TGFbSignaling_Marisa.png"),
  height = 5,
  width = 10,
  res = 400,
  units = "in"
)
ComplexHeatmap::Heatmap(
  corData,
  cluster_columns = T,
  cluster_rows = F,
  col = viridis::cividis(100),
  column_title = "Correlation of Res and Exh scores with \ndifferent signature scores in Marisa data",
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", corData[i, j]), x, y, gp = gpar(fontsize = 10))
  }, show_heatmap_legend = T, name = "Pearson's\ncorrelation\ncoefficient\n"
)
dev.off()
```

## Scores vs Annotation


### Boxplots stratified by MMR, CMS and tumour location
MSI status in the MArisa et al paper was analyzed using a panel of five different microsatellite loci from the Bethesda reference panel. MSI-high tumors were further classified as deficient MMR (dMMR), and both MSI-low and MSS tumors as proficient MMR (pMMR).
These reproduces Figure 4A and 4B (right panels) and Suppl Figure 8 (right panel).
```{r, fig.height = 5, fig.width = 6}
library(ggsignif)

scoreColumns <- which(colnames(annMar)  %in% scoreNames)

annMarLong <- annMar %>%
  pivot_longer(
    cols = scoreColumns[1]:scoreColumns[length(scoreColumns)],
    names_to = "Signature",
    values_to = "Score"
  ) %>%
  data.frame(check.rows = F, check.names = F)


##---- MMR
colnames(annMarLong)[colnames(annMarLong) == "mmr.status.ch1"] <- "MMR"

pdf(
  paste0(figPath, "Marisa_Boxplot_Scores_MMR_Bulk.pdf"),
  height = 6,
  width = 5.2
)
annMarLong %>%
  filter(Sample_Type == "Cancer") %>%
  filter(MMR != "N/A") %>%
  filter(grepl("Bulk", Signature)) %>%
  filter(! grepl("CD8NK", Signature) & ! grepl("All", Signature)) %>%
  ggplot(., aes(x = MMR, y = Score, color = MMR)) +
  geom_boxplot() +
  geom_signif(
    comparisons = list(c("dMMR", "pMMR")),
    map_signif_level = TRUE,
    tip_length = 0,
    vjust = 1.4,
    col = "gray40"
  ) +
  facet_wrap( ~ Signature, scales = "free", ncol = 2) +
  scale_y_continuous(breaks = equal_breaks(n = nBreak, s = scalingFactor),
                     expand = c(scalingFactor, 0)) +
  scale_color_manual(values =  rev(c(brewer.pal(9, "Purples") [c(4, 6, 9)]))) +
  currentTheme +
    theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
        axis.ticks = element_blank(),
    legend.position = "bottom"
    )
dev.off()

##------- CMS
cmsCol <- c(
  brewer.pal(11, "BrBG")[3],
  brewer.pal(11, "PRGn")[2],
  brewer.pal(9, "Blues")[5],
  brewer.pal(11, "BrBG")[10],
  "grey50"
)

pdf(
  paste0(figPath, "Marisa_Boxplot_Scores_CMS_Bulk.pdf"),
  height = 6,
  width = 5.5
)
annMarLong %>%
  filter(complete.cases(CMS)) %>%
  filter(grepl("Bulk", Signature)) %>%
  filter(! grepl("CD8NK", Signature) & ! grepl("All", Signature)) %>%
  ggplot(., aes(x = CMS, y = Score, color = CMS)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap( ~ Signature, scales = "free", ncol = 2) +
  scale_y_continuous(breaks = equal_breaks(n = nBreak, s = scalingFactor),
                     expand = c(scalingFactor, 0)) +
  scale_color_manual(values = cmsCol) +
  currentTheme +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
        axis.ticks = element_blank(),
    legend.position = "bottom"
    )
dev.off()


annMarLong$tumor.location.ch1[
  annMarLong$tumor.location.ch1 == "distal"
] <- "Distal"

annMarLong$tumor.location.ch1[
  annMarLong$tumor.location.ch1 == "proximal"
] <- "Proximal"

pdf(
  paste0(figPath, "Marisa_Boxplot_Scores_Location_Cancer_Bulk.pdf"),
  height = 6,
  width = 5
)
annMarLong %>%
  filter(Sample_Type == "Cancer") %>%
  filter(grepl("Bulk", Signature)) %>%
  filter(!grepl("CD8NK", Signature) & !grepl("All", Signature)) %>%
  ggplot(.,
         aes(x = tumor.location.ch1, y = Score, color = tumor.location.ch1)) +
  geom_boxplot(outlier.shape = NA) +
  ggpubr::geom_signif(
    comparisons = list(c("Distal", "Proximal")),
    map_signif_level = TRUE,
    tip_length = 0,
    vjust = 1.4,
    col = "gray40"
  ) +
  labs(colour = "Location") +
  facet_wrap(~ Signature, scales = "free", ncol = 2) +
  scale_y_continuous(breaks = equal_breaks(n = nBreak, s = scalingFactor),
                     expand = c(scalingFactor, 0)) +
scale_color_manual(values = brewer.pal(8, "Greys")[c(5, 7)]) +
  currentTheme +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "bottom"
  )
dev.off()

```

#### p values of CMS boxplots
```{r}
my_comparisons <- list(c("CMS1", "CMS2"), 
                       c("CMS1", "CMS3"), 
                       c("CMS1", "CMS4"),
                       c("CMS1", "N"),
                       c("CMS2", "CMS3"),
                       c("CMS2", "CMS4"),
                       c("CMS2", "N"),
                       c("CMS3", "CMS4"),
                       c("CMS3", "N"),
                       c("CMS4", "N")
                       )


dd <- annMarLong %>%
  filter(complete.cases(CMS)) %>%
  filter(grepl("Bulk", Signature)) %>%
  filter(! grepl("CD8NK", Signature) & ! grepl("All", Signature)) %>%
  dplyr::select(CMS, Signature, Score) %>% 
  data.frame()
  

signatureNames <- unique(dd$Signature)


ll <- sapply(signatureNames, function(y) {
  d <- dd %>%
    filter(Signature == y) %>%
    data.frame()
  
  sapply(my_comparisons, function(x) {
    comp1 <- x[1]
    comp2 <- x[2]
    
    d <- d %>%
      filter(CMS == comp1 | CMS == comp2) %>%
      data.frame()
    
    res <-
      wilcox.test(d$Score[d$CMS == comp1], d$Score[d$CMS == comp2])$p.value
    return(res)
    
  })
})

row.names(ll) <- sapply(my_comparisons, function(x) paste(x[1], x[2], sep ="_"))


## replace non-significant items with NA
ll[ll > 0.05] <- NA


tiff(
  paste0(figPath, "CMS_pvals_Marisa.tiff"),
  height = 3.3,
  width = 5,
  units = "in",
  res = 300
)
ComplexHeatmap::Heatmap(t(ll), 
                        col = viridis::viridis(200), 
                        name = "p-value\n", 
                        na_col = "gray", 
                        cluster_rows = F, 
                        cluster_columns = F, 
                        column_title = "Marisa\n Wilcox test p-values")
dev.off()

```


# OS
```{r}
# 561 samples
library(survival)
timeCol = "OS.time"
eventCol = "OS"
sn <- "OS"

annMarFilt <- annMarFilt[complete.cases(annMarFilt$Age), ]
annMarFilt <- annMarFilt[complete.cases(annMarFilt[, timeCol]), ]
annMarFilt$MMR[annMarFilt$MMR == "N/A"] <- NA
```
 
## Lasso for Cox to choose covariates
In order to decide which of the clinical factors to include, we perform lasso on the Cox models. We then use the coef of the lambda.min to decide which ones to include or exclude from the model. Based on the results below, we decide include 3 variables CMS, Age and Stage, as they all had non-zero lambda.min.but we do not include MMR.
```{r}
library(tidyverse)
library(broom)
library(glmnet)

set.seed(20201214)
annMarFilt_Sel <-
  annMarFilt[, c(
    "MMR",
    "CMS",
    "Age",
    "Stage",
    "OS.time",
    "OS"
  )]

## 442 patients after removing NAs
annMarFilt_Sel <- annMarFilt_Sel[complete.cases(annMarFilt_Sel) &
annMarFilt_Sel$OS.time > 0,]

sur <- Surv(
  time = as.numeric(annMarFilt_Sel[, timeCol]), 
  event = as.numeric(annMarFilt_Sel[, eventCol])) 


pred <- annMarFilt_Sel[, 
                          ! colnames(annMarFilt_Sel)  %in% c(timeCol, eventCol)]

x <- model.matrix(~ MMR + CMS + Age + Stage, pred)


y <- as.matrix(annMarFilt_Sel[, c(timeCol, eventCol)])
colnames(y) <- c("time", "status")

fit <- glmnet(
  x = x, 
  y = y, 
  family = "cox",
  alpha = 1)

plot(fit, label = T)
cv.fit <- cv.glmnet(x, y, family = "cox", alpha = 1)
plot(cv.fit)


coef(cv.fit, s = "lambda.min")


coxph(Surv(as.numeric(annMarFilt_Sel[, timeCol]), as.numeric(annMarFilt_Sel[, eventCol])) ~
        Age + Stage + MMR + CMS,
      data = annMarFilt_Sel)

coxph(Surv(as.numeric(annMarFilt_Sel[, timeCol]), as.numeric(annMarFilt_Sel[, eventCol])) ~
        Age + Stage + CMS,
      data = annMarFilt_Sel)

```

## Generate different format of the data
### Wide, 2/3 group scores/gene
```{r}

scoreColumns <- which(colnames(annMarFilt) %in% scoreNames)

wide2 <- data.frame(
  ## keep annotations and stage and age
  annMarFilt[, - c(scoreColumns)],
  ## add two group scores and genes
  as.data.frame(sapply(c(scoreNames), function(x) {
  case_when(
    annMarFilt[, x] >= median(annMarFilt[, x], na.rm = T) ~ "High",
    annMarFilt[, x] < median(annMarFilt[, x], na.rm = T) ~ "Low"
  )
})), check.names = F
)


wide3 <- data.frame(## keep annotations and stage and age
  annMarFilt[, - c(scoreColumns)],
  ## add two group scores and genes
  as.data.frame(sapply(c(scoreNames), function(x) {
    case_when(
      annMarFilt[, x] >= quantile(annMarFilt[, x], probs = 0.65) ~ "High",
      annMarFilt[, x] <= quantile(annMarFilt[, x], probs = 0.35) ~ "Low",
      
      annMarFilt[, x] > quantile(annMarFilt[, x], probs = 0.35) &
        annMarFilt[, x] < quantile(annMarFilt[, x], probs = 0.65)  ~ "Intermediate"
    )
  })) , check.names = F)
```

### Long 2/3 groups score/gene
```{r}
scoreAnnot <- pivot_longer(annMarFilt, 
                           cols = scoreColumns[1]:scoreColumns[length(scoreColumns)], 
                           values_to = "Score",
                           names_to = "Signature") %>% 
  data.frame()


scoreAnnot <- scoreAnnot %>%
  data.frame() %>%
  group_by(Signature) %>%
  # mutate(MedianScore = median(TotalScore)) %>%
  mutate(
    MedianScore = median(Score),
    Quantile_low = quantile(Score, probs = 0.35),
    Quantile_High = quantile(Score, probs = 0.65),
    Score_2status = case_when(
      Score >= MedianScore ~ "High score",
      Score < MedianScore ~ "Low score"
    ),
    Score_3status = case_when(
      Score >= Quantile_High ~ "High score",
      Score <= Quantile_low ~ "Low score",
      Score < Quantile_High &
        Score > Quantile_low ~ "Intermediate score"
    )
  ) %>%
  data.frame()

```

```{r}
covarColumns <- c("Age", "Stage", "CMS")
```

### OS: Score + covariates

* Significant signature scores associated with OS when considering 2 groups: NK_Exh_Bulk 
* Significant signature scores associated with OS when considering 3 groups: DDR, NK_Exh_Bulk, TP53_signaling" "CD4_Res_Bulk"


```{r}

cox_OS_Score_2groups <-
  coxph_multitest(
    data = scoreAnnot,
    variableNameCol = "Signature",
    variableCol = "Score_2status",
    timeCol = timeCol,
    eventCol = eventCol,
    nStrata = 2,
    returnCovarStat = F,
    covarCol = covarColumns,
    nCores = 10
  )


cox_OS_Score_2groups_sig <- cox_OS_Score_2groups %>% filter(`Pr(>|z|)` < 0.05) %>% 
  data.frame(check.names = F)

cox_OS_Score_2groups_sig$VariableName




cox_OS_Score_3groups <-
  coxph_multitest(
    data = scoreAnnot,
    variableNameCol = "Signature",
    variableCol = "Score_3status",
    timeCol = timeCol,
    eventCol = eventCol,
    nStrata = 3,
    returnCovarStat = F,
    covarCol = covarColumns,
    nCores = 10
  )

cox_OS_Score_3groups_sig <- cox_OS_Score_3groups %>% filter(`Pr(>|z|)` < 0.05) %>% 
  data.frame(check.names = F)

cox_OS_Score_3groups_sig$VariableName

```


Generate the survival plots for the significant signatures ( "NK_Exh_Bulk") - related to Suppl Figure 13.
```{r}
 
pdf(
  paste0(figPath, "OS_Marisa_Score2groups_Cox.pdf"),
  height = 4,
  width = 3.5
)
for (s in unique(cox_OS_Score_2groups_sig$VariableName)) {
  print(
    survival_plot(
      data =
        eMar[, rownames(annMarFilt)],
      # stratify = "expr",
      stratify = "score",
      annot = annMarFilt,
      scoreCol =  s,
      gene = NULL,
      covariate = NULL,
      isCategoricalCov = TRUE,
      timeCol = paste0(sn, ".time"),
      eventCol = sn,
      nGroup = 2,
      confInt = F,
      ylabel = paste0(sn),
      cols =
        rev(brewer.pal(10, "Paired"))
    )
  )
}
dev.off()

```

### OS: Score Pairs + covariate
```{r}

trmSigs <- c(
  "CD8_Res_Bulk",
  "CD4_Res_Bulk",
  "NK_Res_Bulk",
  "CD8NK_Res_Bulk"
  )


exhSigs <- c(
  "CD8_Exh_Bulk",
  "CD4_Exh_Bulk",
  "NK_Exh_Bulk",
  "CD8NK_Exh_Bulk"
  )


canSigs <-
  c(
    "OXPHOS",
    "Peroxisome",
    "Glycolysis",
    "ReactiveOx",
    "Hypoxia",
    
    "DNA_Repair",
    "Antigen_Presentation",
    
    "HPX_EGF_EMT",
    "TGFbEMT",
    "EMT",
    "Mes_Thiery",
    "Epi_Thiery",
    
    "IL6_JAK_STAT3",
    "TGFb_Signaling", 
    "TP53_signaling",
    "WNT_signaling",
    "KRAS_signaling"
  )

exhTGFbCan <- c(exhSigs, canSigs)
trmTGFbCan <- c(trmSigs, canSigs)


wide2Comb <- wide2

for(s in trmSigs){
  for(e in exhTGFbCan){
    currentColumn <- paste(s, e, sep = "--")
    wide2Comb[, currentColumn] <- paste(wide2Comb[, s], s , "&",
                                      wide2Comb[, e], e)
  }
} 
  
for(s in exhSigs){
  for(e in trmTGFbCan){
    currentColumn <- paste(s, e, sep = "--")
    wide2Comb[, currentColumn] <- paste(wide2Comb[, s], s , "&",
                                      wide2Comb[, e], e)
  }
}  



wide2Comb[, "CD8_Res_Bulk--CD4_Res_Bulk"] <-
  paste(wide2Comb[, "CD8_Res_Bulk"],
        "CD8_Res_Bulk" ,
        "&",
        wide2Comb[, "CD4_Res_Bulk"],
        "CD4_Res_Bulk")
 
wide2Comb[, "CD8_Res_Bulk--NK_Res_Bulk"] <-
  paste(wide2Comb[, "CD8_Res_Bulk"],
        "CD8_Res_Bulk" ,
        "&",
        wide2Comb[, "NK_Res_Bulk"],
        "NK_Res_Bulk")
 
wide2Comb[, "CD4_Res_Bulk--NK_Res_Bulk"] <-
  paste(wide2Comb[, "CD4_Res_Bulk"],
        "CD4_Res_Bulk" ,
        "&",
        wide2Comb[, "NK_Res_Bulk"],
        "NK_Res_Bulk")

 
wide2Comb[, "CD8_Exh_Bulk--CD4_Exh_Bulk"] <-
  paste(wide2Comb[, "CD8_Exh_Bulk"],
        "CD8_Exh_Bulk" ,
        "&",
        wide2Comb[, "CD4_Exh_Bulk"],
        "CD4_Exh_Bulk")
 
wide2Comb[, "CD8_Exh_Bulk--NK_Exh_Bulk"] <-
  paste(wide2Comb[, "CD8_Exh_Bulk"],
        "CD8_Exh_Bulk" ,
        "&",
        wide2Comb[, "NK_Exh_Bulk"],
        "NK_Exh_Bulk")
 
wide2Comb[, "CD4_Exh_Bulk--NK_Exh_Bulk"] <-
  paste(wide2Comb[, "CD4_Exh_Bulk"],
        "CD4_Exh_Bulk" ,
        "&",
        wide2Comb[, "NK_Exh_Bulk"],
        "NK_Exh_Bulk")

 
wide2CombLong <- wide2Comb %>%
  # pivot_longer(., cols = 152:ncol(wide2Comb),
  pivot_longer(., cols = 148:ncol(wide2Comb),
               values_to = "Comb_2status", names_to = "Comb_Name") %>%
  data.frame(check.names = F)


saveRDS(wide2CombLong, paste0(outPath, "Marisa_Survival_OS_ScoreExpr_wide_Str.RDS"))

# wide2CombLong <- readRDS(paste0(outPath, "Marisa_Survival_OS_ScoreExpr_wide_Str.RDS"))
```


## Pairwise comparisons with covariates

 
```{r}
# fit <- survfit(Surv(OS.time, OS) ~ Comb_2status, data = wide2CombLongSel2)
# 
# wide2CombLongSel2$OS.time <- as.numeric(wide2CombLongSel2$OS.time)
# wide2CombLongSel2$OS <- as.numeric(wide2CombLongSel2$OS)
```
  
```{r}

##------- IF NOT USING THE SELECTED COMBINATION ABOVE:
wide2CombLong$Sig1 <- sapply(wide2CombLong$Comb_Name, function(x){
  unlist(strsplit(x, "--"))[1]
})

wide2CombLong$Sig2 <- sapply(wide2CombLong$Comb_Name, function(x){
  unlist(strsplit(x, "--"))[2]
})


cox_OS_2Comb_Extremes6 <- data.frame(matrix(ncol = 17))
colnames(cox_OS_2Comb_Extremes6) <- c(
  "variable",
  "coef",
  "exp(coef)",
  "se(coef)",
  "z",
  "Pr(>|z|)",
  "exp(-coef)",
  "lower .95",
  "upper .95",
  "C",
  "se(C)",
  "test",
  "df",
  "pvalue",
  "formula",
  "VariableName", 
  "Comparison"
)

for(i in unique(wide2CombLong$Comb_Name)) {
  wide2CombLongSel <- wide2CombLong[wide2CombLong$Comb_Name == i, ]
  combns <- combn(unique(wide2CombLongSel$Comb_2status), 2)
  
  for (j in 1:ncol(combns)) {
    wide2CombLongSel2 <-
      wide2CombLongSel[wide2CombLongSel$Comb_2status %in% combns[, j], ]
    
    cur_cox_OS_2Comb_Extremes6 <-
      coxph_multitest(
        data = wide2CombLongSel2,
        variableNameCol = "Comb_Name",
        variableCol = "Comb_2status",
        timeCol = timeCol,
        eventCol = eventCol,
        nStrata = 2,
        returnCovarStat = F,
        covarCol = covarColumns,
        nCores = 10
      )

    cur_cox_OS_2Comb_Extremes6$Comparison <- paste(combns[1, j],
                                                   combns[2, j], sep = " vs ")
    
    cox_OS_2Comb_Extremes6 <-
      rbind(cox_OS_2Comb_Extremes6, cur_cox_OS_2Comb_Extremes6)
  }
  
}

## remove the first row of NA:
cox_OS_2Comb_Extremes6 <- cox_OS_2Comb_Extremes6[complete.cases(cox_OS_2Comb_Extremes6$variable), ]
## 1320   17 -- 220 pairs

  
cox_OS_2Comb_Extremes6_sig <-
  cox_OS_2Comb_Extremes6 %>% filter(`Pr(>|z|)` < 0.05) %>%
  data.frame(check.names = F) 
  



cox_OS_2Comb_Extremes6_sig$Group1 <- sapply(
  cox_OS_2Comb_Extremes6_sig$Comparison,
  function(x) unlist(strsplit(x, " vs "))[1])

cox_OS_2Comb_Extremes6_sig$Group2 <- sapply(
  cox_OS_2Comb_Extremes6_sig$Comparison,
  function(x) unlist(strsplit(x, " vs "))[2])
       

## add a column for the contrast
for(i in 1:nrow(cox_OS_2Comb_Extremes6_sig)) {
  curVariable <- cox_OS_2Comb_Extremes6_sig$variable[i]
  curGroup1 <- cox_OS_2Comb_Extremes6_sig$Group1[i]
  curGroup2 <- cox_OS_2Comb_Extremes6_sig$Group2[i]
  
  if (curVariable == curGroup1) {
    cox_OS_2Comb_Extremes6_sig$Contrast[i] <-
      paste0(curVariable, " - ", curGroup2)
  } else if (curVariable == curGroup2) {
    cox_OS_2Comb_Extremes6_sig$Contrast[i] <-
      paste0(curVariable, " - ", curGroup1)
  }
}  


write.csv(cox_OS_2Comb_Extremes6_sig, 
          paste0(outPath, "Marisa_Cox_OS_ExtremeScorePairs_significant_Str_pairwise6.csv"), 
          row.names = F)
```

```{r}

cox_OS_2Comb_Extremes6_sig$Comparison <- gsub("High", "H", cox_OS_2Comb_Extremes6_sig$Comparison)
cox_OS_2Comb_Extremes6_sig$Comparison <- gsub("Low", "L", cox_OS_2Comb_Extremes6_sig$Comparison)

cox_OS_2Comb_Extremes6_sig$Contrast <- gsub("High", "H", cox_OS_2Comb_Extremes6_sig$Contrast)
cox_OS_2Comb_Extremes6_sig$Contrast <- gsub("Low", "L", cox_OS_2Comb_Extremes6_sig$Contrast)



pdf(
  paste0(figPath, "OS_Marisa_ExtremeScorePairs_Cox_Str_pairwise.pdf"),
  height = 7,
  width = 8
)

for (cc in unique(cox_OS_2Comb_Extremes6_sig$VariableName)) {
  
  currentCox <- cox_OS_2Comb_Extremes6_sig[cox_OS_2Comb_Extremes6_sig$VariableName == cc, ]
  
  annot <-
    wide2CombLong[wide2CombLong$Comb_Name == cc , ]
  
  
  
  tt <- data.frame(table(annot[, "Comb_2status"]))
  tt$Var1 <- as.character(tt$Var1)
  tt$Freq <- as.character(tt$Freq)
  
  
  for (i in 1:nrow(tt)) {
    annot$Comb_2status_n[annot[, "Comb_2status"] == tt$Var1[i]] <-
      paste0(tt$Var1[i], " (", tt$Freq[i], ")")
  }
  
  annot <- annot[, c("Comb_2status_n", timeCol, eventCol)]
  
  
  annot[, timeCol] <- as.numeric(annot[, timeCol])
  annot[, eventCol] <- as.numeric(annot[, eventCol])
  
  fitValues <- survfit(Surv(time = annot[, timeCol],
                            event = annot[, eventCol]) ~
                         annot$Comb_2status_n)

  
  signifCompr <- currentCox$Contrast
  
  print(survminer::ggsurvplot (
    fitValues,
    data = annot,
    fun = "pct",
    pval = T,
    conf.int = F,
    palette = cols,
    xlab = "Time",
    legend.title = "",
    title = "Marisa",
    # subtitle = paste0("p-value (Cox)", pValCox, ";", " N = ", nSamples, " in each group \n "),
    subtitle = paste0("Significant contrasts:\n ", paste(signifCompr, collapse = "\n")), 
    # legend.labs = c("High CD4_Exh & Low NK_Exh", "Low CD4_Exh & High NK_Exh"),
    legend = c(.4, .2),
    font.x = 20,
    font.y = 20,
    font.tickslab = 18,
    font.main = 20,
    font.submain = c(12, "italic"),
    font.legend = 11

  ))

}
dev.off()


```




# RFS

```{r}

# 559 samples
library(survival)
timeCol = "RFS.time"
eventCol = "RFS"
sn <- "RFS"

annMarFilt <- annMarFilt[complete.cases(annMarFilt$Age), ]
annMarFilt <- annMarFilt[complete.cases(annMarFilt[, timeCol]), ]
annMarFilt$MMR[annMarFilt$MMR == "N/A"] <- NA
```
 
 
Remove samples that have time = 0; these are mostly stage IV samples that have recurrence (event = 1) and few stage II and III samples that have or do not have recurrence. In total 38 samples. We keep the rest of the stage 4 samples (30).
```{r}
# annMarFilt[(annMarFilt$RFS == 1 &
#                   annMarFilt$RFS.time == 0), 
#                c("Stage", "RFS.time", "RFS")]

## 521 samples
annMarFilt <- annMarFilt[
  ! annMarFilt$RFS.time == 0, 
]

```

## Lasso for Cox to choose covariates
In order to decide which of the clinical factors to include, we need to perform lasso on the Cox models. We then use the coef of the lambda.min to decide which ones to include or exclude from the model. Based on the results below, we need to include all 4 variables.
```{r}
library(tidyverse)
library(broom)
library(glmnet)

set.seed(20201214)
annMarFilt_Sel <-
  annMarFilt[, c(
    "MMR",
    "CMS",
    "Age",
    "Stage",
    "RFS.time",
    "RFS"
  )]

## 381 patients after removing NAs
annMarFilt_Sel <- annMarFilt_Sel[complete.cases(annMarFilt_Sel) &
annMarFilt_Sel$RFS.time > 0,]

sur <- Surv(
  time = as.numeric(annMarFilt_Sel[, timeCol]), 
  event = as.numeric(annMarFilt_Sel[, eventCol])) 


pred <- annMarFilt_Sel[, 
                          ! colnames(annMarFilt_Sel)  %in% c(timeCol, eventCol)]

x <- model.matrix(~ MMR + CMS + Age + Stage, pred)


y <- as.matrix(annMarFilt_Sel[, c(timeCol, eventCol)])
colnames(y) <- c("time", "status")

fit <- glmnet(
  x = x, 
  y = y, 
  family = "cox",
  alpha = 1)

plot(fit, label = T)
cv.fit <- cv.glmnet(x, y, family = "cox", alpha = 1)
plot(cv.fit)


coef(cv.fit, s = "lambda.min")

```

## Generate different format of the data
### Wide, 2/3 group scores/gene
```{r}
scoreColumns <- which(colnames(annMarFilt) %in% scoreNames)

wide2 <- data.frame(
  ## keep annotations and stage and age
  annMarFilt[, - c(scoreColumns)],
  ## add two group scores and genes
  as.data.frame(sapply(c(scoreNames), function(x) {
  case_when(
    annMarFilt[, x] >= median(annMarFilt[, x], na.rm = T) ~ "High",
    annMarFilt[, x] < median(annMarFilt[, x], na.rm = T) ~ "Low"
  )
})), check.names = F
)


wide3 <- data.frame(## keep annotations and stage and age
  annMarFilt[, - c(scoreColumns)],
  ## add two group scores and genes
  as.data.frame(sapply(c(scoreNames), function(x) {
    case_when(
      annMarFilt[, x] >= quantile(annMarFilt[, x], probs = 0.65) ~ "High",
      annMarFilt[, x] <= quantile(annMarFilt[, x], probs = 0.35) ~ "Low",
      
      annMarFilt[, x] > quantile(annMarFilt[, x], probs = 0.35) &
        annMarFilt[, x] < quantile(annMarFilt[, x], probs = 0.65)  ~ "Intermediate"
    )
  })) , check.names = F)
```

### Long 2/3 groups score/gene
```{r}
scoreAnnot <- pivot_longer(annMarFilt, 
                           cols = scoreColumns[1]:scoreColumns[length(scoreColumns)], 
                           values_to = "Score",
                           names_to = "Signature") %>% 
  data.frame()


scoreAnnot <- scoreAnnot %>%
  data.frame() %>%
  group_by(Signature) %>%
  # mutate(MedianScore = median(TotalScore)) %>%
  mutate(
    MedianScore = median(Score),
    Quantile_low = quantile(Score, probs = 0.35),
    Quantile_High = quantile(Score, probs = 0.65),
    Score_2status = case_when(
      Score >= MedianScore ~ "High score",
      Score < MedianScore ~ "Low score"
    ),
    Score_3status = case_when(
      Score >= Quantile_High ~ "High score",
      Score <= Quantile_low ~ "Low score",
      Score < Quantile_High &
        Score > Quantile_low ~ "Intermediate score"
    )
  ) %>%
  data.frame()

```

```{r}
covarColumns <- c("Age", "Stage", "CMS", "MMR")
```


### RFS: Score + covariates
* Significant signature scores associated with PFI when considering 2 groups: "Epi_Thiery"  "HPX_EGF_EMT" "Peroxisome"  "TP53_signaling"
* Significant signature scores associated with PFI when considering 3 groups:"HPX_EGF_EMT"  "Hypoxia"    "Peroxisome"  "TP53_signaling" "KRAS_signaling


```{r}


cox_RFS_Score_2groups <-
  coxph_multitest(
    data = scoreAnnot,
    variableNameCol = "Signature",
    variableCol = "Score_2status",
    timeCol = timeCol,
    eventCol = eventCol,
    nStrata = 2,
    returnCovarStat = F,
    covarCol = covarColumns,
    nCores = 10
  )


cox_RFS_Score_2groups_sig <- cox_RFS_Score_2groups %>% filter(`Pr(>|z|)` < 0.05) %>% 
  data.frame(check.names = F)

cox_RFS_Score_2groups_sig$VariableName


cox_RFS_Score_3groups <-
  coxph_multitest(
    data = scoreAnnot,
    variableNameCol = "Signature",
    variableCol = "Score_3status",
    timeCol = timeCol,
    eventCol = eventCol,
    nStrata = 3,
    returnCovarStat = F,
    covarCol = covarColumns,
    nCores = 10
  )

cox_RFS_Score_3groups_sig <- cox_RFS_Score_3groups %>% filter(`Pr(>|z|)` < 0.05) %>% 
  data.frame(check.names = F)

cox_RFS_Score_3groups_sig$VariableName


```


### RFS: Score Pairs + covariate
```{r}


trmSigs <- c(
  "CD8_Res_Bulk",
  "CD4_Res_Bulk",
  "NK_Res_Bulk",
  "CD8NK_Res_Bulk"
  )


exhSigs <- c(
  "CD8_Exh_Bulk",
  "CD4_Exh_Bulk",
  "NK_Exh_Bulk",
  "CD8NK_Exh_Bulk"
  )



# canSigs <-
#   c(
#     "DDR",
#     "OXPHOS",
#     "TP53_signaling",
#     "HPX_EGF_EMT",
#     "Peroxisome",
#     "Mes_Thiery",
#     "Epi_Thiery",
#     "IL6_JAK_STAT3",
#     "TGFb_Signaling"
#   )

canSigs <-
  c(
    "OXPHOS",
    "Peroxisome",
    "Glycolysis",
    "ReactiveOx",
    "Hypoxia",
    
    "DNA_Repair",
    "Antigen_Presentation",
    
    "HPX_EGF_EMT",
    "TGFbEMT",
    "EMT",
    "Mes_Thiery",
    "Epi_Thiery",
    
    "IL6_JAK_STAT3",
    "TGFb_Signaling", 
    "TP53_signaling",
    "WNT_signaling",
    "KRAS_signaling"
  )

exhTGFbCan <- c(exhSigs, canSigs)
trmTGFbCan <- c(trmSigs, canSigs)


wide2Comb <- wide2

for(s in trmSigs){
  for(e in exhTGFbCan){
    currentColumn <- paste(s, e, sep = "--")
    wide2Comb[, currentColumn] <- paste(wide2Comb[, s], s , "&",
                                      wide2Comb[, e], e)
  }
} 
  
for(s in exhSigs){
  for(e in trmTGFbCan){
    currentColumn <- paste(s, e, sep = "--")
    wide2Comb[, currentColumn] <- paste(wide2Comb[, s], s , "&",
                                      wide2Comb[, e], e)
  }
}  


wide2Comb[, "CD8_Res_Bulk--CD4_Res_Bulk"] <-
  paste(wide2Comb[, "CD8_Res_Bulk"],
        "CD8_Res_Bulk" ,
        "&",
        wide2Comb[, "CD4_Res_Bulk"],
        "CD4_Res_Bulk")
 
wide2Comb[, "CD8_Res_Bulk--NK_Res_Bulk"] <-
  paste(wide2Comb[, "CD8_Res_Bulk"],
        "CD8_Res_Bulk" ,
        "&",
        wide2Comb[, "NK_Res_Bulk"],
        "NK_Res_Bulk")
 
wide2Comb[, "CD4_Res_Bulk--NK_Res_Bulk"] <-
  paste(wide2Comb[, "CD4_Res_Bulk"],
        "CD4_Res_Bulk" ,
        "&",
        wide2Comb[, "NK_Res_Bulk"],
        "NK_Res_Bulk")

 
wide2Comb[, "CD8_Exh_Bulk--CD4_Exh_Bulk"] <-
  paste(wide2Comb[, "CD8_Exh_Bulk"],
        "CD8_Exh_Bulk" ,
        "&",
        wide2Comb[, "CD4_Exh_Bulk"],
        "CD4_Exh_Bulk")
 
wide2Comb[, "CD8_Exh_Bulk--NK_Exh_Bulk"] <-
  paste(wide2Comb[, "CD8_Exh_Bulk"],
        "CD8_Exh_Bulk" ,
        "&",
        wide2Comb[, "NK_Exh_Bulk"],
        "NK_Exh_Bulk")
 
wide2Comb[, "CD4_Exh_Bulk--NK_Exh_Bulk"] <-
  paste(wide2Comb[, "CD4_Exh_Bulk"],
        "CD4_Exh_Bulk" ,
        "&",
        wide2Comb[, "NK_Exh_Bulk"],
        "NK_Exh_Bulk")

 


wide2CombLong <- wide2Comb %>%
  pivot_longer(., cols = 148:ncol(wide2Comb),
               values_to = "Comb_2status", names_to = "Comb_Name") %>%
  data.frame(check.names = F)


saveRDS(wide2CombLong, paste0(outPath, "Marisa_Survival_RFS_ScoreExpr_wide_Str.RDS"))

# wide2CombLong <- readRDS(paste0(outPath, "Marisa_Survival_RFS_ScoreExpr_wide_Str.RDS"))


```
## Pairwise comparisons with covariates
```{r}

##------- IF NOT USING THE SELECTED COMBINATION ABOVE:
wide2CombLong$Sig1 <- sapply(wide2CombLong$Comb_Name, function(x){
  unlist(strsplit(x, "--"))[1]
})

wide2CombLong$Sig2 <- sapply(wide2CombLong$Comb_Name, function(x){
  unlist(strsplit(x, "--"))[2]
})


cox_PFI_2Comb_Extremes6 <- data.frame(matrix(ncol = 17))
colnames(cox_PFI_2Comb_Extremes6) <- c(
  "variable",
  "coef",
  "exp(coef)",
  "se(coef)",
  "z",
  "Pr(>|z|)",
  "exp(-coef)",
  "lower .95",
  "upper .95",
  "C",
  "se(C)",
  "test",
  "df",
  "pvalue",
  "formula",
  "VariableName", 
  "Comparison"
)

for(i in unique(wide2CombLong$Comb_Name)) {
  wide2CombLongSel <- wide2CombLong[wide2CombLong$Comb_Name == i, ]
  combns <- combn(unique(wide2CombLongSel$Comb_2status), 2)
  
  for (j in 1:ncol(combns)) {
    # print(combns[, j])
    
    wide2CombLongSel2 <-
      wide2CombLongSel[wide2CombLongSel$Comb_2status %in% combns[, j], ]
    
    cur_cox_PFI_2Comb_Extremes6 <-
      coxph_multitest(
        data = wide2CombLongSel2,
        variableNameCol = "Comb_Name",
        variableCol = "Comb_2status",
        timeCol = timeCol,
        eventCol = eventCol,
        nStrata = 2,
        returnCovarStat = F,
        covarCol = covarColumns,
        nCores = 10
      )
    
    cur_cox_PFI_2Comb_Extremes6$Comparison <- paste(combns[1, j],
                                                   combns[2, j], sep = " vs ")
    
    cox_PFI_2Comb_Extremes6 <-
      rbind(cox_PFI_2Comb_Extremes6, cur_cox_PFI_2Comb_Extremes6)
  }
  
}

## remove the first row of NA:
cox_PFI_2Comb_Extremes6 <- cox_PFI_2Comb_Extremes6[complete.cases(cox_PFI_2Comb_Extremes6$variable), ]
## 1320   17 -- 220 pairs

  
cox_PFI_2Comb_Extremes6_sig <-
  cox_PFI_2Comb_Extremes6 %>% filter(`Pr(>|z|)` < 0.05) %>%
  data.frame(check.names = F) 
  


cox_PFI_2Comb_Extremes6_sig$Group1 <- sapply(
  cox_PFI_2Comb_Extremes6_sig$Comparison,
  function(x) unlist(strsplit(x, " vs "))[1])

cox_PFI_2Comb_Extremes6_sig$Group2 <- sapply(
  cox_PFI_2Comb_Extremes6_sig$Comparison,
  function(x) unlist(strsplit(x, " vs "))[2])
       

## add a column for the contrast
for(i in 1:nrow(cox_PFI_2Comb_Extremes6_sig)) {
  curVariable <- cox_PFI_2Comb_Extremes6_sig$variable[i]
  curGroup1 <- cox_PFI_2Comb_Extremes6_sig$Group1[i]
  curGroup2 <- cox_PFI_2Comb_Extremes6_sig$Group2[i]
  
  if (curVariable == curGroup1) {
    cox_PFI_2Comb_Extremes6_sig$Contrast[i] <-
      paste0(curVariable, " - ", curGroup2)
  } else if (curVariable == curGroup2) {
    cox_PFI_2Comb_Extremes6_sig$Contrast[i] <-
      paste0(curVariable, " - ", curGroup1)
  }
}  



write.csv(cox_PFI_2Comb_Extremes6_sig, 
          paste0(outPath, "Marisa_Cox_RFS_ExtremeScorePairs_significant_Str_pairwise6.csv"), 
          row.names = F)
```

```{r}

cox_PFI_2Comb_Extremes6_sig$Comparison <- gsub("High", "H", cox_PFI_2Comb_Extremes6_sig$Comparison)
cox_PFI_2Comb_Extremes6_sig$Comparison <- gsub("Low", "L", cox_PFI_2Comb_Extremes6_sig$Comparison)

cox_PFI_2Comb_Extremes6_sig$Contrast <- gsub("High", "H", cox_PFI_2Comb_Extremes6_sig$Contrast)
cox_PFI_2Comb_Extremes6_sig$Contrast <- gsub("Low", "L", cox_PFI_2Comb_Extremes6_sig$Contrast)

pdf(
  paste0(figPath, "RFS_Marisa_ExtremeScorePairs_Cox_Str_pairwise.pdf"),
  height = 6.5,
  width = 8
)

for (cc in unique(cox_PFI_2Comb_Extremes6_sig$VariableName)) {
  
  currentCox <- cox_PFI_2Comb_Extremes6_sig[cox_PFI_2Comb_Extremes6_sig$VariableName == cc, ]
  
  annot <-
    wide2CombLong[wide2CombLong$Comb_Name == cc, ]
 
  
  tt <- data.frame(table(annot[, "Comb_2status"]))
  tt$Var1 <- as.character(tt$Var1)
  tt$Freq <- as.character(tt$Freq)
  
  
  for (i in 1:nrow(tt)) {
    annot$Comb_2status_n[annot[, "Comb_2status"] == tt$Var1[i]] <-
      paste0(tt$Var1[i], " (", tt$Freq[i], ")")
  }
  
  annot <- annot[, c("Comb_2status_n", timeCol, eventCol)]
  
  
  annot[, timeCol] <- as.numeric(annot[, timeCol])
  annot[, eventCol] <- as.numeric(annot[, eventCol])
  
  fitValues <- survfit(Surv(time = annot[, timeCol],
                            event = annot[, eventCol]) ~
                         annot$Comb_2status_n)

  
  signifCompr <- currentCox$Contrast
  
  print(survminer::ggsurvplot (
    fitValues,
    data = annot,
    fun = "pct",
    pval = T,
    conf.int = F,
    palette = cols,
    xlab = "Time",
    legend.title = "",
    title = "Marisa",
    subtitle = paste0("Significant contrasts:\n ", paste(signifCompr, collapse = "\n")), 
    legend = c(.4, .2),
    font.x = 20,
    font.y = 20,
    font.tickslab = 18,
    font.main = 20,
    font.submain = c(12, "italic"),
    font.legend = 11

  ))

}
dev.off()


```
 
 




### Selected scatterplots
**Scatterplot of CD8NK_Exh_Bulk vs TGFb_Signaling, coloured by CMS subtypes (Figure 6B, right scatterplot)**
```{r, fig.height = 4.7, fig.width = 4.3}
x <- "CD8NK_Exh_Bulk"
y <- "TGFb_Signaling"


d <- annMar %>%
  filter(! annMar$Sample_Type  %in%  "Normal") %>%
  mutate(medIm = median(!!sym(x)),
         medOt = median(!!sym(y)))

palette <- cmsCol
# palette <- rev(c(brewer.pal(9, "Purples") [c( 6, 9)]))
pal <- list(values=palette, na.value="gray80")

corVal <- round(cor.test(d[, x], d[, y])[[4]], 2)

print(
  d %>%
    ggplot(.,
           # aes_string(x, y, color = "MMR")) +
           aes_string(x, y, color = "CMS")) +
    geom_point(alpha = 0.8) +
    invoke(scale_colour_manual, pal) +
    geom_vline(xintercept = d$medIm) +
    geom_hline(yintercept = d$medOt) +
    scale_x_continuous(
                   breaks = equal_breaks(n = nBreak, s = scalingFactor),
                   expand = c(scalingFactor, 0)
                 ) +
    scale_y_continuous(
                   breaks = equal_breaks(n = nBreak, s = scalingFactor),
                   expand = c(scalingFactor, 0)
                 ) +
    ggtitle(paste0("Marisa (cor = ", corVal, ")")) +
    currentTheme +
    guides(color = guide_legend(nrow=2))
)
```


**Scatterplot of CD8NK_Exh_Bulk vs NK_Res_Bulk, coloured by CMS subtypes (Figure 7B, right scatterplot)**
```{r, fig.height = 4.7, fig.width = 4.3}
x <- "CD8NK_Exh_Bulk"
y <- "NK_Res_Bulk"

d <- annMar %>%
  filter(! annMar$Sample_Type  %in%  "Normal") %>%
  mutate(medIm = median(!!sym(x)),
         medOt = median(!!sym(y)))

palette <- cmsCol
# palette <- rev(c(brewer.pal(9, "Purples") [c( 6, 9)]))
pal <- list(values=palette, na.value="gray80")

corVal <- round(cor.test(d[, x], d[, y])[[4]], 2)

print(
  d %>%
    ggplot(.,
           # aes_string(x, y, color = "MMR")) +
           aes_string(x, y, color = "CMS")) +
    geom_point(alpha = 0.8) +
    invoke(scale_colour_manual, pal) +
    geom_vline(xintercept = d$medIm) +
    geom_hline(yintercept = d$medOt) +
    scale_x_continuous(
                   breaks = equal_breaks(n = nBreak, s = scalingFactor),
                   expand = c(scalingFactor, 0)
                 ) +
    scale_y_continuous(
                   breaks = equal_breaks(n = nBreak, s = scalingFactor),
                   expand = c(scalingFactor, 0)
                 ) +
    ggtitle(paste0("Marisa (cor = ", corVal, ")")) +
    currentTheme +
    guides(color = guide_legend(nrow=2))
)
```


**Scatterplot of NK_Res_Bulk vs NK_Exh_Bulk, coloured by CMS subtypes (Figure 8B, right scatterplot)**
```{r, fig.height = 4.7, fig.width = 4.3}
x <- "NK_Res_Bulk"
y <- "NK_Exh_Bulk"

d <- annMar %>%
  filter(! annMar$Sample_Type  %in%  "Normal") %>%
  mutate(medIm = median(!!sym(x)),
         medOt = median(!!sym(y)))

palette <- cmsCol
# palette <- rev(c(brewer.pal(9, "Purples") [c( 6, 9)]))
pal <- list(values=palette, na.value="gray80")

corVal <- round(cor.test(d[, x], d[, y])[[4]], 2)

print(
  d %>%
    ggplot(.,
           # aes_string(x, y, color = "MMR")) +
           aes_string(x, y, color = "CMS")) +
    geom_point(alpha = 0.8) +
    invoke(scale_colour_manual, pal) +
    geom_vline(xintercept = d$medIm) +
    geom_hline(yintercept = d$medOt) +
    scale_x_continuous(
                   breaks = equal_breaks(n = nBreak, s = scalingFactor),
                   expand = c(scalingFactor, 0)
                 ) +
    scale_y_continuous(
                   breaks = equal_breaks(n = nBreak, s = scalingFactor),
                   expand = c(scalingFactor, 0)
                 ) +
    ggtitle(paste0("Marisa (cor = ", corVal, ")")) +
    currentTheme +
    guides(color = guide_legend(nrow=2))
)
```

     
# Mutation
There are mutation info available for the BRAF, KRAS, and TP53 in Marisa et al data; here we generate boxplots of Res/Exh scores stratified by mutation status.
```{r, fig.height = 7.5, fig.width = 13}

annMarLong2 <- annMarLong %>%
  pivot_longer(
    cols = c(BRAF, TP53, KRAS),
    values_to = "Mut",
    names_to = "Genes"
  ) %>%
  filter(grepl("Bulk", Signature))

annMarLong2$Gene_Signature <- paste(annMarLong2$Genes, annMarLong2$Signature, sep = "_")

annMarLong2$Mut[annMarLong2$Mut == "M"] <- "Mut"
annMarLong2$Mut <- factor(annMarLong2$Mut, levels = c("WT", "Mut"))

currentcombs <- unique(annMarLong2$Gene_Signature)[order(unique(annMarLong2$Gene_Signature))]

p <- list()

for(i in 1:length(currentcombs)) {
  
  g <- unique(annMarLong2$Gene_Signature)[i]
  
  currentAnnMarLong2 <- annMarLong2 %>% 
    filter(Gene_Signature == g)

  ylab <- unique(currentAnnMarLong2$Signature)
  currentGene <- as.character(unique(currentAnnMarLong2$Genes))
  title <- currentGene
  
  p[[g]] <- currentAnnMarLong2 %>%
    filter(Sample_Type == "Cancer" &
             Mut != "N/A") %>%
    ggplot(data = ., aes(x = Mut, y = Score, color = Mut)) +
    geom_boxplot() +
    geom_jitter(position = position_jitter(0.2), alpha = 0.4) +
    ylab(ylab) +
  geom_signif(comparisons = list(c("Mut", "WT")),
              map_signif_level = TRUE, tip_length = 0, vjust= 1.4) +
    ggtitle(title) +
    scale_color_manual(values = brewer.pal(4, "Dark2")[c(1, 3)]) +
    scale_y_continuous(
      breaks = equal_breaks(n = nBreak, s = scalingFactor),
      expand = c(scalingFactor, 0)
    ) +
    theme_bw() +
    theme(axis.title.x = element_blank(), legend.position = "none")
}



library(gridExtra)
p <- p[currentcombs]

pdf(
  paste0(figPath, "Boxplot_Marisa_Muts_ScoresAllExhTrm_3Genes_Str.pdf"),
  height = 7.5,
  width = 13)
grid.arrange(
  grobs = p,
  ncol = 10
)
dev.off()

```

      
# Session information
Here are the list of tools and packages (along with their versions) used in this document. 
```{r}
sessionInfo()
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 598px;"></div>
    
          
          
          
          



















