---
title: 'Residency and exhaustion signatures for CD4 T cells in Zhang et al Smart-seq2 single cell data'
author: 
  - name: 'Sepideh Foroutan'
    affiliation: 'Huntington Cancer Immunotherapy Lab'
    url: https://www.monash.edu/discovery-institute/huntington-lab
date: '10-06-2020'
output:
  rmdformats::readthedown:
    fig_width: 12
    fig_height: 6
    gallery: TRUE
    highlight: tango
    lightbox: TRUE
    self_contained: TRUE
    thumbnails: FALSE
    number_sections: TRUE	
    toc_depth: 3
    use_bookdown: TRUE
    code_folding: hide
  html_document2:
    df_print: paged
params:
  update_date: !r paste("Last updated on:", Sys.Date() )
editor_options: 
  chunk_output_type: inline
---
`r params$update_date`

<style>
body {
text-align: justify}
</style>


```{r knitr_init, echo=FALSE, results="asis", cache=FALSE}

library(knitr)
library(rmdformats)

options(max.print = "75")
opts_chunk$set(
  # echo = FALSE,
  cache = TRUE,
  # prompt = FALSE,
  # tidy = FALSE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  # results = FALSE,
  root.dir = normalizePath("."))
opts_knit$set(width = 65)

```

# Set up and overview
```{r file set up, echo=F, results=F, message=F, warning=F}

library(DT)
library(Seurat)
library(SingleCellExperiment)
library(singscore)
library(slingshot)
library(tidyverse)
library(RColorBrewer)

mainDir <- getwd()
outPath <- "../output/NK/"
figPath <- "../figure/NK/"
dataPath <- "../data/"
scriptPath <- "../script/"


tpmPath <- "../data/GSE146771_Zhang_CRC/"
hpaGoUniprotPath <- "../data/HPA_GO_Uniprot/"
ccleLcmPath <- "../data/CCLE_LCM/"

ifelse(!dir.exists(file.path(mainDir, outPath)), dir.create(file.path(mainDir, outPath)), FALSE)
ifelse(!dir.exists(file.path(mainDir, figPath)), dir.create(file.path(mainDir, figPath)), FALSE)
ifelse(!dir.exists(file.path(mainDir, dataPath)), dir.create(file.path(mainDir, dataPath)), FALSE)
ifelse(!dir.exists(file.path(mainDir, scriptPath)), dir.create(file.path(mainDir, scriptPath)), FALSE)


options(digits = 3)

equal_breaks <- function(n = nBreak, s = scalingFactor, ...){
  function(x){
    # rescaling
    d <- s * diff(range(x)) / (1+2*s)
    round( seq(min(x)+d, max(x)-d, length=n), 2)
  }
}

nBreak = 3
scalingFactor = 0.05

textSize <- 1.2

currentTheme <- theme_minimal() +
  theme(
    # panel.background = element_blank()
    axis.title = element_text(size = rel(textSize)),
    axis.text = element_text(angle = 0, size = rel(textSize)),
    # strip.background = element_rect(colour = "#f0f0f0", fill = "#f0f0f0"),
    strip.text = element_text(size = rel(textSize)),
    axis.line = element_line(colour = "black", size = 0.5),
    legend.position = 'top',
    # legend.title = element_text(size = rel(textSize), face = "italic"),
    # legend.text = element_text(size = rel(textSize)),
    legend.key.size = unit(1, 'lines'),
    ## increase the line space
    plot.title = element_text(
      face = "bold",
      size = rel(textSize),
      hjust = 0.5
    )
  )

cols <-  c(
  brewer.pal(8, "Dark2")[-5],
  brewer.pal(10, "Paired"),
  brewer.pal(12, "Set3"),
  brewer.pal(9, "Blues")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Oranges")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Greens")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Purples")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Reds")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "Greys")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "BuGn")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "PuRd")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "BuPu")[c(8, 3, 7, 4, 6, 9, 5)],
  brewer.pal(9, "YlGn")[c(8, 3, 7, 4, 6, 9, 5)])

source(paste0(scriptPath, "visReduction_function.R"))
```


# Read SMART-seq2 CD4 data
We read the SMART-seq2 data, and generate a Seurat object for it. The data is already in TPM format.
Note that among 18 patients in Zhang et al study, 3 of them are MSI-H (P0123, P0413 and P0825), and the other 15 patients are MSS. So we add these info to the annotation file and calculate the percent of mitochondrial genes.

```{r, cache = T}
## dim:15179 10469

crcSmart <-
   data.table::fread(
    paste0(tpmPath, "GSE146771_CRC.Leukocyte.Smart-seq2.TPM.txt")
  )

rnames <- crcSmart$V1
crcSmart <- crcSmart[, 2:ncol(crcSmart)]
rownames(crcSmart) <- rnames

crcSmart <- as.matrix(crcSmart)
rownames(crcSmart) <- rnames


metaSmart <-
  read.table(
    paste0(tpmPath, "GSE146771_CRC.Leukocyte.Smart-seq2.Metadata.txt"),
    header = T,
    sep = "\t"
  )

metaSmart$Sub_Cluster <- as.character(metaSmart$Sub_Cluster)
metaSmart$Sub_Cluster <- sapply(metaSmart$Sub_Cluster, function(x){
  unlist(strsplit(x, "_"))[2]
})


##----- assign colours to cells
immCells <- metaSmart$Sub_Cluster[grepl("CD8", metaSmart$Sub_Cluster)|
                                    grepl("NK", metaSmart$Sub_Cluster) |
                                    grepl("CD4", metaSmart$Sub_Cluster)|
                                    grepl("ILC3", metaSmart$Sub_Cluster)]

cellCols <- cols[1:length(unique(immCells))]
names(cellCols) <- unique(immCells)[order(unique(immCells))]
nkCols <- cellCols[grepl("NK-", names(cellCols))]

##----- generate Seurat object

subSmartNK <- crcSmart[, grepl('NK-', metaSmart$Sub_Cluster) ]
subMetaSmartNK <- metaSmart[grepl('NK-', metaSmart$Sub_Cluster), ]


# all(subMetaSmartNK$CellName == colnames(subSmartNK))
rownames(subMetaSmartNK) <- subMetaSmartNK$CellName

subMetaSmartNK$Sub_Cluster <- as.character(subMetaSmartNK$Sub_Cluster)

sdRow <- matrixStats::rowSds(subSmartNK)
subSmartNK <- subSmartNK[ sdRow!=0 , ]

nk <-
  CreateSeuratObject(
    counts = subSmartNK,
    project = 'Zhang_SmartSeq_NK',
    meta.data = subMetaSmartNK,
    min.cells = 5
  )

nk[['percent.mt']] <- PercentageFeatureSet(nk, pattern = '^MT-')

nk$MSI <- "MSS"
nk$MSI[nk$Sample %in% c("P0123", "P0413", "P0825")] <- "MSI-H"

## save the RDS file of the NK
# saveRDS(nk, paste0(dataPath, "GSE146771_Zhang_CRC/NK_Zhang_Smartseq_Seurat.RDS"))
```


## Read GPCRs and TFs
In order to annotate genes based on whether or not they are TFs, GPCR or associated with GPCR signalling, we use the human protein atlas and get the GPCR and TF annotations from there. For GPCR, we also downloaded the GPCR related genes from QuickGO and selected the genes annotated as gpcr in uniprot too. 
```{r}
hpa <- read.delim(paste0(hpaGoUniprotPath, "proteinatlas.tsv"), header = T)

GPCRs <-
  hpa[grepl("coupled", hpa$Gene.description, ignore.case = T) |
        grepl("coupled", hpa$Protein.class, ignore.case = T), c(
          "Gene",
          "Ensembl",
          "Protein.class",
          "Gene.description",
          "Uniprot" ,
          "Blood.RNA...NK.cell..NX."
        )]

GPCRs <- GPCRs[order(GPCRs$Blood.RNA...NK.cell..NX., decreasing = T), ]


TFs <- hpa[grepl("Transcription factor", hpa$Gene.description, ignore.case = T) |
        grepl("Transcription factor", hpa$Protein.class, ignore.case = T), c(
          "Gene",
          "Ensembl",
          "Protein.class",
          "Gene.description",
          "Uniprot" ,
          "Blood.RNA...NK.cell..NX."
        )]
  
TFs <- TFs[order(TFs$Blood.RNA...NK.cell..NX., decreasing = T), ]
 
  
go <- read.delim(paste0(hpaGoUniprotPath, "QuickGO-annotations-1592213989958-20200615.tsv"), header = T)

go <- unique(go$SYMBOL)

## 780 genes common between these
# length(intersect(GPCRs$Gene, go))

## 2035 unique genes
gpr <- unique(c(as.character(GPCRs$Gene), as.character(go)))
gpr <- as.character(gpr)
gpr <- gpr[! grepl("Homo sapiens", gpr)]
gpr <- gpr[! grepl("human", gpr)]
## MGRF

uniprot <- read.csv(paste0(hpaGoUniprotPath, "7tmrlist_uniprot.csv"), 
                    stringsAsFactors = F)

uniprot <- uniprot[grepl("HUMAN", uniprot$genes), ]
uniprot <- sapply(uniprot$genes, function(x){ unlist(strsplit(x, "_"))[1]})

uniprot <- as.character(uniprot)


tars <- read.csv(paste0(hpaGoUniprotPath, "targets_and_families.csv"), 
                 stringsAsFactors = F)

tars <- tars$HGNC.symbol[tars$Type == "gpcr"]
tars <- tars[! tars == ""]
tars <- tars[!duplicated(tars)]

gpr <- unique(c(gpr, uniprot, tars))

```


# Data analysis and visualisations

## Initial Res and Exh signatures

### Define genes for NK analysis
Here we have put together all the signature from T and NK cell residency/exhaustion/dysfunction that we collected from the literature (see Table 1 in the paper). Note that as the residency and exhaustion program is less known in NK cells, we included the genes in both T and NK cells from the literature when studying NK cells. These are 1744 genes, of which 1525 genes exist in the single cell data. These include some well known canonnical Res and Exh markers.

```{r}

TcellSigsList <-
  readRDS(paste0(
    dataPath,
    "Signatures_literature/Tcell_signature_list_literature.RDS"
  ))

nkcellSigsList <- 
    readRDS(paste0(
    dataPath,
    "Signatures_literature/NKcell_signature_list_literature.RDS"
  ))

## 1744 genes
selGenes <- unique(c(unlist(TcellSigsList), unlist(nkcellSigsList)))
```


# NK analysis
```{r}

subSmartNK <- as.matrix(GetAssayData(nk))

## 1525 genes
nkExpr <- subSmartNK[rownames(subSmartNK) %in% selGenes, ]

corMat <- cor(t(nkExpr), method = "spearman")


## function for flattening the cor matrix
flattenCorrMatrix <- function(cormat
) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut]
  )
}

corStat <- flattenCorrMatrix(corMat)

```



Canonical Res and Exh markers for CD8:

* Res: "CD69", "ITGAE", "ITGA1", "RGS1", "ZNF683"
* Exh: "HAVCR2", "PDCD1", "CTLA4", "TIGIT", "LAG3"

Regarding the Res genes, we select genes that have correlation more than 0.35 with canonical Res markers. This step results in 75 genes.

Regarding the Exh genes, we select genes that have correlation more than 0.25 with canonical Exh markers. We get one gene based on their correlation with CTLA4 ( DUSP4), and one gene based on the correlation with PDCD1 (SIRPG). Most genes comes from HAVCR2, TIGIT and LAG3. Note that HAVCR2 and LAG3 have also high expression in blood NK cells. This step results in 56 genes.


As some genes (N = 31) are common between both Res and Exh genes obtained above, i.e. have correlations with both Res and Exh canonical markers, we remove them from both gene lists.

```{r}
trmMinGenes <- c("CD69", "ITGAE", "ITGA1", "RGS1", "ZNF683")
exhMinGenes <- c("HAVCR2", "PDCD1", "CTLA4", "TIGIT", "LAG3")
```


```{r}
trmCors <- corStat %>% filter(cor > 0.35 & (row %in% trmMinGenes | column %in% trmMinGenes))
trmGenes <- unique(c(as.character(trmCors$row),
                     as.character(trmCors$column)))



texCors <- corStat %>% filter(cor > 0.25 & (row %in% exhMinGenes | column %in% exhMinGenes))
texGenes <- unique(c(as.character(texCors$row),
                     as.character(texCors$column)))

```


Initial Res genes, before removing genes common with Exh genes:
```{r}
trmGenes
```

Initial Exh genes, before removing genes common with Res genes:
```{r}
texGenes
```

Genes common between the Res and Exh gene lists:
```{r}
comGs <- intersect(texGenes, trmGenes)
comGs

trmGenes <- trmGenes[! trmGenes %in% comGs]
texGenes <- texGenes[! texGenes %in% comGs]
```


### Sparse canonical correlation
Here, we also perform sparse canonical correlation (CCA) to remove genes that cause correlations between Res and Exh gene sets, although these genes are not common between the two signatures. We visualise the cross-correlation between the Res and Exh genes before and after removing genes based on sparse CCA.

Based on the results of the CCA, we remove "GEM" "ID3" "IRF4"  "ITGAE" "LDLRAD4" "CAPG"  from the initial Res gene list and "FABP5"  "GAPDH"  "ENTPD1" from the initial Exh gene list.
```{r, fig.height=10, fig.width = 12, warning = F, message = F, cache = T,  fig.cap = "Cross-correlation between all initial Res and Exh signature genes after CCA and removal of genes causing high correlation between the two signatures"}
library(PMA)

cca_nk_init <- CCA(
  t(subSmartNK[rownames(subSmartNK) %in% trmGenes, ]), 
   t(subSmartNK[rownames(subSmartNK) %in% texGenes, ]))

trmCorGenes <- colnames(t(subSmartNK[rownames(subSmartNK) %in% trmGenes, ]))[which(cca_nk_init$u[, 1] <= -0.1)]
# [1] "GEM" "ID3" "IRF4"  "ITGAE"  "LDLRAD4"  "CAPG"   
texCorGenes <- colnames(t(subSmartNK[rownames(subSmartNK) %in% texGenes, ]))[which(cca_nk_init$v[, 1] <= -0.1)]
# "FABP5"  "GAPDH"  "ENTPD1"

##---- remove genes that have non-zero values in CCA:

trmGenes <- trmGenes[! trmGenes %in% trmCorGenes]
texGenes <- texGenes[! texGenes %in% texCorGenes]


nkExpr <- subSmartNK[unique(c(trmGenes, texGenes)), ]

corMat <- cor(t(nkExpr), method = "spearman")


markerCols <- c("gray90", "gray20")
names(markerCols) <- c("Res", "Exh")

sampleAnnHm <-
  ComplexHeatmap::HeatmapAnnotation(Marker = as.character(c(rep(
    "Res", length(trmGenes)
  ), rep(
    "Exh", length(texGenes)
  ))),
  col = list(Marker = markerCols))

# pdf(paste0(figPath, "CrossCor_ZhangSmart_NK_ResTexGenes_rmComGenes_rmCCA_20200928.pdf"), height = 13, width = 14)
ComplexHeatmap::Heatmap(corMat,
                        show_column_names = T, 
                        top_annotation = sampleAnnHm, 
                        col = circlize::colorRamp2(
                          c(min(corMat), 0, max(corMat)), c("yellow3", "white", "navy")), 
                        # col = viridis::plasma(100), 
                        name = "Rho")
# dev.off()
```

Remove genes that are common across both: "KIR2DL4" "DUSP4"   "ACP5" 
```{r}
saveRDS(list(trmGenes = trmGenes, texGenes = texGenes), 
        paste0(outPath, "Initial_Res_Exh_Genes_NK_Corr_20200928.RDS"))

# trmExhGenes <- readRDS(paste0(outPath, "Initial_Res_Exh_Genes_NK_Corr_20200928.RDS"))
# 
# trmGenes <- trmExhGenes$trmGenes
# texGenes <- trmExhGenes$texGenes
```


## Standard Seurat pipeline
```{r, warning = F, message = F, cache = T}
nk <- FindVariableFeatures(nk, selection.method = "vst", nfeatures = 2000)

all.genes <- rownames(nk)
nk <- ScaleData(nk, features = all.genes)

nk <- RunPCA(nk, features = VariableFeatures(object = nk), verbose = F)

# DimHeatmap(nk, dims = 1:30, cells = 500, balanced = TRUE)
# ElbowPlot(nk, ndims = 40)

nk <- FindNeighbors(nk, dims = 1:10)
nk <- FindClusters(nk, resolution = 0.5, random.seed = 20200928)

nk <- RunUMAP(nk, dims = 1:10, n.neighbors = 30, min.dist = 0.3, seed.use = 20200928)
```


```{r, fig.height = 4, fig.width = 9, fig.cap = "Relationship between percent mito genes or number of detected genes and library size (from Seurat)"}

plot1 <-
  FeatureScatter(nk,
                 feature1 = "nCount_RNA",
                 feature2 = "percent.mt",
                 group.by = "Sub_Cluster", 
                 cols = nkCols)
plot2 <-
  FeatureScatter(nk,
                 feature1 = "nCount_RNA",
                 feature2 = "nFeature_RNA",
                 group.by = "Sample", 
                 cols = cols)
plot1 + plot2

```


```{r, fig.height=8, fig.width = 4, fig.cap = "Number of features, library size and percent mito genes in NK cells across patients"}
VlnPlot(
  nk,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  ncol = 1,
  group.by = "Sample",
  pt.size = 0, cols = brewer.pal(9, "Set1")
)
```

### Highly Variable genes
```{r, fig.cap = "Highly variable genes"}
# nk <- FindVariableFeatures(nk, selection.method = "vst", nfeatures = 2000)

top10 <- head(VariableFeatures(nk), 20)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(nk)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```

```{r, fig.height = 6, fig.width = 6, fig.cap = "Gene loadings for PC1 and PC2"}
VizDimLoadings(nk, dims = 1:2, reduction = "pca")
```



## Score SMART-seq2 CD4 T cells based on initial Res and Exh signatures
```{r, message = F, warning = F}
rd <- rankGenes(as.matrix(GetAssayData(nk, slot = "data", assay = "RNA")))
     
nk$CuratedTrmGenes <- simpleScore(rankData = rd, 
                              upSet = trmGenes, 
                              knownDirection = T, 
                              centerScore = F, 
                              dispersionFun = var)$TotalScore    
     
nk$CuratedExhGenes <- simpleScore(rankData = rd, 
                              upSet = texGenes, 
                              knownDirection = T, 
                              centerScore = F, 
                              dispersionFun = var)$TotalScore    
## Save data after scores
# saveRDS(nk, file = paste0(outPath, "DontPush/Zhang_SmartSeq_SeuratObj_NK_20200928_UMAP_scores.RDS"))

```


### Scatterplot of scores (related to Figure 1)
We generate scatterplots of the Initial Res vs Exh scores. In addition to reproducing a part of Figure 1, we colour cells based on different available annotations.

```{r, fig.height = 5, fig.width = 5, fig.cap = "Scatter plot of scores comparing initial Residency and initial Exhaustion signature scores"}

cell_attrsNK <- nk[[]]

highTrmPct <- quantile(cell_attrsNK$CuratedTrmGenes, probs = 0.70)
highExhPct <- quantile(cell_attrsNK$CuratedExhGenes, probs = 0.70)

# pdf(
#   paste0(
#     figPath,
#     "ScatterPlot_ZhangSmart_NK_CuratedTRM_TEX_AllCells_lines_20200928.pdf"
#   ),
#   height = 4.3,
#   width = 4.85
# )

cell_attrsNK %>%
  ggplot(.,
         aes(x = CuratedTrmGenes, y = CuratedExhGenes)) +
  geom_point(alpha = 0.8, size = 1, color = "gray50") +
  geom_hline(yintercept = highExhPct,  col = "gray10", size = 1) +
  geom_vline(xintercept = highTrmPct, col = "gray10", size = 1) +
  xlab("Initial residency score") +
  ylab("Initial exhaustion score") +
  currentTheme +
  guides(color = guide_legend(nrow = 3,
                              override.aes = list(size = 3)))


cell_attrsNK %>%
  ggplot(.,
         aes(x = CuratedTrmGenes, y = CuratedExhGenes, color = Sub_Cluster)) +
  geom_point(alpha = 0.8, size = 1) +
  scale_color_manual(values = nkCols) +
  geom_hline(yintercept = highExhPct,  col = "gray10", size = 1) +
  geom_vline(xintercept = highTrmPct, col = "gray10", size = 1) +
  xlab("Initial residency score") +
  ylab("Initial exhaustion score") +
  currentTheme +
  guides(color = guide_legend(nrow = 3,
                              override.aes = list(size = 3)))

cell_attrsNK %>%
  ggplot(., aes(x = CuratedTrmGenes, y = CuratedExhGenes, color = Tissue)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_hline(yintercept = highExhPct,  col = "gray10", size = 1) +
  geom_vline(xintercept = highTrmPct, col = "gray10", size = 1) +
  scale_color_manual(values = cols) +
    xlab("Initial residency score") +
  ylab("Initial exhaustion score") +
  currentTheme +
  guides(color = guide_legend(nrow = 3,
                              override.aes = list(size = 3)))
 
 
cell_attrsNK %>%
  ggplot(., aes(x = CuratedTrmGenes, y = CuratedExhGenes, color = MSI)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_hline(yintercept = highExhPct,  col = "gray10", size = 1) +
  geom_vline(xintercept = highTrmPct, col = "gray10", size = 1) +
  scale_color_manual(values = cols) +
    xlab("Initial residency score") +
  ylab("Initial exhaustion score") +
  currentTheme +
  guides(color = guide_legend(nrow = 3,
                              override.aes = list(size = 3)))
# dev.off()
```


## Markers for NK cells
```{r}
##----------- In both Tum and normal, after removing blood cells
cell_attrsNK_Tissue <-
  cell_attrsNK %>% 
  filter(Tissue != "P")

cell_attrsNK_Tissue$TissueExhCells[cell_attrsNK_Tissue$CuratedExhGenes >
                                      highExhPct &
                                      cell_attrsNK_Tissue$CuratedTrmGenes <
                                      highTrmPct] <-
  "NK_Exh"


cell_attrsNK_Tissue$TissueExhCells[cell_attrsNK_Tissue$CuratedTrmGenes >
                                      highTrmPct &
                                      cell_attrsNK_Tissue$CuratedExhGenes < highExhPct] <-
  "NK_Res"

cell_attrsNK_Tissue$TissueExhCells[is.na(cell_attrsNK_Tissue$TissueExhCells)] <- "Other"

table(cell_attrsNK_Tissue$TissueExhCells)

# save cell names for these groups: use these in percentile comparisons

NK_Exh_Cells <- as.character(
  cell_attrsNK_Tissue$CellName[
    cell_attrsNK_Tissue$TissueExhCells == "NK_Exh"])


NK_Res_Cells <- as.character(
  cell_attrsNK_Tissue$CellName[
    cell_attrsNK_Tissue$TissueExhCells == "NK_Res"])
```

Subset the Seurat object to also remove peripheral blood cells.
```{r}
NKTissue <- nk[, ! nk$Tissue == "P" ]
# all(cell_attrsNK_Tissue$CellName == colnames(NKTissue))
NKTissue@meta.data <- cell_attrsNK_Tissue
NKTissue <- SetIdent(NKTissue, value = "TissueExhCells")
```

### NK Exh markers
```{r, cache = T}
##------- Find markers for NK-Exh vs NK-Res-nonEx
markers_NKExh <-
  FindMarkers(
    NKTissue,
    ident.1 = "NK_Exh",
    ident.2 = "NK_Res",
    only.pos = TRUE,
    min.pct = 0.3,
    logfc.threshold = 0.3
  )

markers_NKExh$gene <- rownames(markers_NKExh)
markers_NKExh$Marker <- "NK_Exh"

## 51 genes
DEGs_NKExh <- markers_NKExh[markers_NKExh$avg_logFC > 0.3 & markers_NKExh$p_val_adj < 0.05, ]

# dim(DEGs_NKExh)
##----------- DE using MAST
# BiocManager::install("MAST")
# library(MAST)
markers_NKExh_mast <-
  FindMarkers(
    NKTissue,
    ident.1 = "NK_Exh",
    ident.2 = "NK_Res",
    only.pos = TRUE,
    min.pct = 0.3,
    logfc.threshold = 0.3,
    test.use = "MAST"
  )

markers_NKExh_mast$gene <- rownames(markers_NKExh_mast)
markers_NKExh_mast$Marker <- "NK_Exh"

## 45 genes
DEGs_NKExh_mast <- markers_NKExh_mast[markers_NKExh_mast$avg_logFC > 0.3 & markers_NKExh_mast$p_val_adj < 0.05, ]

# dim(DEGs_NKExh_mast)

## 43 common
# length(intersect(DEGs_NKExh_mast$gene, DEGs_NKExh$gene))

## --- Only add genes that came up in MAST and do not exist in 
## Wilcox test

DEGs_NKExh_mast <- DEGs_NKExh_mast[! rownames(DEGs_NKExh_mast) %in% rownames(DEGs_NKExh), ]

##--- 53 genes in total
DEGs_NKExh <- rbind(DEGs_NKExh, DEGs_NKExh_mast)


## Add GPCRs:
DEGs_NKExh$IsGPCR <- FALSE
DEGs_NKExh$IsGPCR[DEGs_NKExh$gene %in% gpr] <- TRUE
## Add TFs:
DEGs_NKExh$IsTF <- FALSE
DEGs_NKExh$IsTF[DEGs_NKExh$gene %in% TFs$Gene] <- TRUE

```


#### Percentile filtration
Filter those high in blood cells and also high in Res cells.
We keep genes whose 75% tile expression in Exh cells are larger than the 90%tile of their expression in blood and Res cells. This results in 19 genes.

```{r}
exprData <- as.matrix(GetAssayData(nk, slot = "data"))

gg <-  DEGs_NKExh$gene
exprData_Genes <- exprData[gg, ]
cell_attrsNK <- nk[[]]


Exh_Pct <-
  rowQuantiles(exprData_Genes[, NK_Exh_Cells], probs = 0.75)

Res_Pct <-
  rowQuantiles(exprData_Genes[, NK_Res_Cells], probs = 0.9)

Blood_Pct <-
  rowQuantiles(exprData_Genes[, cell_attrsNK$Tissue == "P"], probs = 0.9)


pctData_nkExh <- data.frame(cbind(Exh_Pct,
                               Res_Pct, 
                               Blood_Pct
                               ))

pctData_nkExh$gene <- rownames(pctData_nkExh)

pctData_nkExhPass <- pctData_nkExh %>% 
  filter(Exh_Pct > Res_Pct & Exh_Pct> Blood_Pct) %>% 
  data.frame()

pctData_nkExhPass <-
  pctData_nkExhPass[order(pctData_nkExhPass$Exh_Pct, decreasing = T),]

DT::datatable(pctData_nkExhPass, filter = "top") ## 19 genes

DEGs_NKExh$PctPass <- FALSE
DEGs_NKExh$PctPass[DEGs_NKExh$gene %in% pctData_nkExhPass$gene] <- TRUE

# write.table(
#   DEGs_NKExh,
#   paste0(outPath, "Zhang_NKExh_Markers_BasedOnScores_20200928.txt"),
#   row.names = F,
#   sep = "\t"
# )
```


### NK Res markers
```{r}
##------- Find markers for NK-Res vs Nk-Exh
markers_NKRes <-
  FindMarkers(
    NKTissue,
    ident.1 = "NK_Res",
    ident.2 = "NK_Exh",
    only.pos = TRUE,
    min.pct = 0.3,
    logfc.threshold = 0.3
  )

markers_NKRes$gene <- rownames(markers_NKRes)
markers_NKRes$Marker <- "NK_Res"

## 140
DEGs_NKRes <- markers_NKRes[markers_NKRes$avg_logFC > 0.3 & markers_NKRes$p_val_adj < 0.05, ]

# dim(DEGs_NKRes)
##------- use MAST for DE
markers_NKRes_mast <-
  FindMarkers(
    NKTissue,
    ident.1 = "NK_Res",
    ident.2 = "NK_Exh",
    only.pos = TRUE,
    min.pct = 0.3,
    logfc.threshold = 0.3,
    test.use = "MAST"
  )

markers_NKRes_mast$gene <- rownames(markers_NKRes_mast)
markers_NKRes_mast$Marker <- "NK_Res"

## 131 genes
DEGs_NKRes_mast <- markers_NKRes_mast[markers_NKRes_mast$avg_logFC > 0.3 & markers_NKRes_mast$p_val_adj < 0.05, ]

# dim(DEGs_NKRes_mast)

## 118 common
# length(intersect(DEGs_NKRes_mast$gene, DEGs_NKRes$gene))

##  23 genes came up in this method that did not come up in Wilcox!
DEGs_NKRes_mast <- DEGs_NKRes_mast[! rownames(DEGs_NKRes_mast) %in% rownames(DEGs_NKRes), ]

## 81 common genes
##--- 153 genes in total
DEGs_NKRes <- rbind(DEGs_NKRes, DEGs_NKRes_mast)

## Add GPCRs:
DEGs_NKRes$IsGPCR <- FALSE
DEGs_NKRes$IsGPCR[DEGs_NKRes$gene %in% gpr] <- TRUE
## Add TFs:
DEGs_NKRes$IsTF <- FALSE
DEGs_NKRes$IsTF[DEGs_NKRes$gene %in% TFs$Gene] <- TRUE

```


#### Percentile filtration
Filter those high in blood and also high in Exh cells --> 61 genes
```{r}
exprData <- as.matrix(GetAssayData(nk, slot = "data"))

gg <-  DEGs_NKRes$gene
exprData_Genes <- exprData[gg, ]

cell_attrsNK <- nk[[]]


Res_Pct <-
  rowQuantiles(exprData_Genes[, NK_Res_Cells ], probs = 0.65)

Exh_Pct <-
  rowQuantiles(exprData_Genes[, NK_Exh_Cells ], probs = 0.9)

Blood_Pct <-
  rowQuantiles(exprData_Genes[, cell_attrsNK$Tissue == "P"], probs = 0.90)


pctData_nkRes <- data.frame(cbind(Res_Pct,
                               Exh_Pct, 
                               Blood_Pct
                               ))

pctData_nkRes$gene <- rownames(pctData_nkRes)

pctData_nkResPass <- pctData_nkRes %>% 
  filter(Res_Pct > Exh_Pct & Res_Pct > Blood_Pct) %>% 
  data.frame()

pctData_nkResPass <-
  pctData_nkResPass[order(pctData_nkResPass$Res_Pct, decreasing = T),]

DT::datatable(pctData_nkResPass, filter = "top")  ## 61 genes


DEGs_NKRes$PctPass <- FALSE
DEGs_NKRes$PctPass[DEGs_NKRes$gene %in% pctData_nkResPass$gene] <- TRUE

# write.table(
#   DEGs_NKRes,
#   paste0(outPath, "Zhang_NKTrm_Markers_BasedOnScores_20200928.txt"),
#   row.names = F,
#   sep = "\t"
# )
```


## Score SMART-seq2 NK cells using the final Res and Exh signatures
```{r}
library(singscore)
exprData <- as.matrix(GetAssayData(nk, slot = "data"))
rd <- rankGenes(exprData)

NK_Res <- simpleScore(
  rankData = rd,
  upSet =  pctData_nkResPass$gene,
  knownDirection = T,
  centerScore = F
)

NK_Exh <- simpleScore(
  rankData = rd,
  upSet =  pctData_nkExhPass$gene,
  knownDirection = T,
  centerScore = F
)

nk$NK_Res <- NK_Res$TotalScore
nk$NK_Exh <- NK_Exh$TotalScore

```

### Save data after final scores
```{r}
# saveRDS(nk, paste0(outPath, "DontPush/Zhang_SmartSeq_SeuratObj_NK_20200928_UMAP_scores2.RDS"))
# nk <- readRDS(paste0(outPath, "DontPush/Zhang_SmartSeq_SeuratObj_NK_20200928_UMAP_scores2.RDS"))
```

## PCA/UMAP using marker genes
```{r, cache = T}
gg <-  c(pctData_nkExhPass$gene, pctData_nkResPass$gene)

nkTrmExhSig <-
  RunPCA(nk, features = rownames(nk)[rownames(nk) %in% gg], verbose = F)

# ElbowPlot(nkTrmExhSig)

# Clustering and UMAP on marker genes - including both signatures
nkTrmExhSig <- FindNeighbors(nkTrmExhSig, reduction = "pca", dims = 1:10)
nkTrmExhSig <- FindClusters(nkTrmExhSig, resolution = 0.5)

nkTrmExhSig <- RunUMAP(nkTrmExhSig, reduction = "pca", 
          dims = 1:10,  min.dist = 0.2, n.neighbors = 20)

# DimPlot(nkTrmExhSig, reduction = "umap")
# DimPlot(nkTrmExhSig, reduction = "umap", group.by = "Sub_Cluster")
# DimPlot(nkTrmExhSig, reduction = "umap", group.by = "Tissue")
# FeaturePlot(nkTrmExhSig, reduction = "umap", features = "NK_Exh")
# FeaturePlot(nkTrmExhSig, reduction = "umap", features = "NK_Res")
```


# Check expression of genes (realted to Suppl Figure S5 and S6)

```{r, fig.height=8, fig.width=8}
# nk <- readRDS(paste0(outPath, "Zhang_SmartSeq_SeuratObj_NK_20200928_UMAP_scores2.RDS"))

ann <- nk[[]]
exprData <- as.matrix(GetAssayData(nk, slot = "data"))

colnames(ann) <- gsub("NK_Res", "NK_Res", colnames(ann) )

tissueCols <- cols[1:3]
names(tissueCols) <- c("N", "P", "T")


exhColour <- colorRampPalette(brewer.pal(9, 'Blues')[-c(1)])(8)
trmColour <- colorRampPalette(brewer.pal(9, 'Oranges')[-c(1)])(8)

sampleAnnHm <-
  ComplexHeatmap::HeatmapAnnotation(
    df = ann[, c("Sub_Cluster", "Tissue", "NK_Exh", "NK_Res")],
    col = list(Sub_Cluster = nkCols,
               Tissue = tissueCols, 
               NK_Exh = circlize::colorRamp2(seq(min(ann$NK_Exh),
    max(ann$NK_Exh), l = 8), exhColour), 
               NK_Res = circlize::colorRamp2(seq(min(ann$NK_Res),
    max(ann$NK_Res), l = 8), trmColour)
    
    ))


nk_genes <- as.character(quote(
  c(
    ##------- NK receptor genes:
    ## PB CD56bright
    KLRD1,
    ## KLRD1  is the same as CD94
    KLRF1,
    ## KLRF1 is the same as NKp80
    NCAM1,
    ## same as CD56 also in resident NK, not in Kidney
    IL2RA,
    IL7RA,
    KIT,
    ## CD117
    SELL,
    ## CD62L
    NKG2A,
    ## in most resident NKs but not in uterine, and not in CD56 dim
    ITGA5,
    ## CD49e
    CXCR3,
    CCR7,
    NCR1,
    ## NKp46
    
    ## PB CD56dim
    FCGR3A,
    ## CD16
    ##KIRs  also in lung and uterine resident NK
    KIR3DL3,
    KIR2L3,
    KIR2DP1,
    KIR2DL1,
    KIR3DP1,
    KIR2DL4,
    KIR3DL1,
    KIR2DS4,
    KIR3DL2,
    KIR2DS1,
    KIR2DS2,
    KIR2DS3,
    KIR2DS5,
    KIR2DL2,
    KIR2DL5,
    KIR3DS1,
    S1PR5,
    ## S1P5
    CX3CR1,
    CXCR2,
    CXCR1,
    ## PB adaptive NK:
    CD2,
    KLRC2,
    ## NKG2C
    LILRB1,
    ## CD85j
    B3GAT1,
    ## CD57
    ## Bone marrow , SLT,Gut:
    ITGA1,
    ## CD49a,
    ITGAE,
    ## CD103,
    CD69,
    CCR5,
    ## not in lung, uterine and kidney
    CXCR6,
    ## not in lung, uterine and kidney
    ## Uterine:
    CD9,
    ##----- NK secretory proteins
    IFNG,
    ## In PB CD56bright and PB adaptive NK, and not resident NK
    PRF1,
    ## In PB CD56dim
    GZMA,
    GZMB,
    GZMH,
    GZMK,
    GZMM, ## In PB CD56dim)
    ## add PDCD1 and TIGIT:
    PDCD1,
    TIGIT
    
  )
)) [-1]



library(grid)

# pdf(
#   paste0(figPath, "Heatmap_ZhangSmart_NK_selectedGenes_AllCells.pdf"),
#   height = 8,
#   width = 8
# )

ComplexHeatmap::Heatmap(exprData[nk_genes[nk_genes  %in% rownames(exprData)], ],
                        show_column_names = F, 
                        top_annotation = sampleAnnHm, 
                        col = viridis::cividis(100), 
                        row_names_gp = gpar(fontsize = c(13)), 
                        name = "logExpr")


# dev.off()

```

```{r, fig.height = 3.2, fig.width = 8}
gs <-
  c(
    # "IFNG",
    "GZMK",
    "CCR7",
    "SELL",
    "GPR183",
    # "GPR15",
    "PDCD1",
    "CTLA4",
    "HAVCR2",
    "TIGIT")

# pdf(
#   paste0(figPath, "Heatmap_ZhangSmart_NK_selectedFewGenes2_rmIFNG_AllCells_scores.pdf"),
#   height = 3.2,
#   width = 8
# )

ComplexHeatmap::Heatmap(exprData[ gs, ],
                        show_column_names = F, 
                        top_annotation = sampleAnnHm, 
                        col = viridis::cividis(100), 
                        name = "logExpr", 
                        column_title = "NK cells")

# dev.off()
```


## Trajectory analysis
```{r}
## add clusters from markers:
nk$seurat_clusters_markers <- nkTrmExhSig$seurat_clusters

cell_attrsNK <- nk[[]]
exprData <- as.matrix(GetAssayData(nk, slot = "data"))

rd_umap_seurat <- nk@reductions[["umap"]]@cell.embeddings
rd_pca_seurat <- nk@reductions[["pca"]]@cell.embeddings
```

### Generate SCE
```{r}
nk.sce <- as.SingleCellExperiment(nk)

reducedDims(nk.sce) <-
  SimpleList(

    ## from Seurat
    PCA_Seurat = rd_pca_seurat,
    UMAP_Seurat = rd_umap_seurat,
    
    ## scores
    CuratedScores = as.matrix(cell_attrsNK[, c("CuratedTrmGenes", "CuratedExhGenes")]),
    MarkerScores = as.matrix(cell_attrsNK[, c("NK_Res", "NK_Exh")]),
    
    ## PCs from markers
    UMAP_markers = nkTrmExhSig@reductions[["umap"]]@cell.embeddings,
    PCA_markers = nkTrmExhSig@reductions[["pca"]]@cell.embeddings
  )

# saveRDS(nk.sce, paste0(outPath, "DontPush/Zhang_SmartSeq_NK_SingleCellExperiment_20200928.RDS"))
# nk.sce <- readRDS(paste0(outPath, "DontPush/Zhang_SmartSeq_NK_SingleCellExperiment_20200928.RDS"))
```

### Slingshot (realted to Figure 1 in the paper)
```{r, fig.cap= "Trajectory analysis in NK cells using the UMAP and clusters obtained based on our Res and Exh marker genes, coloured by Res and Exh signature scores"}
source(paste0(scriptPath, "slingshot_ggplot.R"))
nk.sce <- slingshot(nk.sce, clusterLabels = 'seurat_clusters_markers', reducedDim = 'UMAP_markers')

sds <- SlingshotDataSet(nk.sce)

textSize <- 1.4

currentThemeTrj <- theme_dark() +
  theme(
    # panel.background = element_blank()
    axis.title = element_text(size = rel(textSize)),
    axis.text = element_text(angle = 0, size = rel(textSize)),
    # strip.background = element_rect(colour = "#f0f0f0", fill = "#f0f0f0"),
    strip.text = element_text(size = rel(textSize)),
    axis.line = element_line(colour = "black", size = 0.5),
    legend.position = 'top',
    legend.title = element_text(size = rel(textSize), face = "italic"),
    legend.text = element_text(size = rel(textSize)),
     legend.key.width = unit(2.4, 'lines'),
    ## increase the line space
    plot.title = element_text(
      face = "bold",
      size = rel(textSize),
      hjust = 0.5
    )
  )

# pdf(
#   paste0(
#     figPath,
#     "Trajectory_ggplot_ZhangSmart_NK_MarkersUMAP_ClusterMarkers_TrmScore.pdf"
#   ),
#   height = 5,
#   width = 8
# )
ggplot() +
  geom_point(size = 1) +
  gg_plot(sds, col = nk.sce$NK_Exh) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  scale_color_viridis_c(
    option = "viridis", 
    name = "NK \nexhaustion\nscore")  +
  currentThemeTrj 

ggplot() +
  geom_point(size = 1) +
  gg_plot(sds, col = nk.sce$NK_Res) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  scale_color_viridis_c(
    option = "viridis", 
    name = "NK \nresidency\nscore")  +
  currentThemeTrj 
# dev.off()

```


Check the expression of some selected genes on the trajectory
```{r, fig.cap="Trajectory analysis in NK cells using the UMAP and clusters obtained based on our Res and Exh marker genes, coloured by the expression of selected genes"}
gs <-
  c("IFNG",
    "GZMK",
    "CCR7",
    "SELL",
    "GPR183",
    "GPR15",
    "PDCD1",
    "CTLA4",
    "HAVCR2",
    "TIGIT")

# pdf(
#   paste0(
#     figPath,
#     "Trajectory_ggplot_ZhangSmart_NK_MarkersUMAP_ClusterMarkers_selGenes.pdf"
#   ),
#   height = 5,
#   width = 8
# )
for(i in gs){
  d0 <- as.matrix(assay(nk.sce))[i,]

print(ggplot() +
  geom_point(size = 1) +
  gg_plot(sds, col = d0) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  scale_color_viridis_c(
    option = "viridis", 
    name = i)  +
  currentThemeTrj )
}
# dev.off()
```



# Filter markers based on the tumour cell data

## CCLE
```{r, cache = T, warning=F, message=F, fig.cap="Expression of NK Res and Exh genes in the CCLE CRC data", fig.height=15, fig.width=8 }
selectedGenes <- c(pctData_nkResPass$gene, pctData_nkExhPass$gene)

dge <- readRDS(paste0(ccleLcmPath, "DGEList_CCLE_panCancer.RDS"))

library(edgeR)
  
kp <- rowSums(cpm(dge) > 1) >= 5 |
  dge$genes$Symbol %in% selectedGenes
summary(kp)

dgeFilt <- dge[kp, ]
dgeFiltNorm <- calcNormFactors(dgeFilt)

logRPKM <- rpkm(dgeFiltNorm, log = T, gene.length = dgeFiltNorm$genes$gene_length)

gg <- selectedGenes[selectedGenes %in% dgeFiltNorm$genes$Symbol]
logRPKM_CRC <- logRPKM[, dgeFiltNorm$samples$type_refined == "colorectal"]

brbg <- rev(RColorBrewer::brewer.pal(11, "PRGn"))
brbg_hcl <- colorspace::diverging_hcl(11,
  h = c(180, 50), c = 80, l = c(20, 95), power = c(0.7, 1.3))

ComplexHeatmap::Heatmap(logRPKM_CRC[gg, ! is.na(colnames(logRPKM_CRC))], 
                        col = brbg_hcl, 
                        show_column_names = F, name = "logRPKM")

```

We subset genes to those that have low expression in CCLE. These are genes whose median expression is =< 0 across samples.
```{r}
ggLogRPKM <- logRPKM_CRC[gg, ! is.na(colnames(logRPKM_CRC))]
markersPassCCLE <- rownames(ggLogRPKM)[rowMedians(ggLogRPKM) <= 0]
markersPassCCLE
```


## LCM (Laser captured micro-dissection)
Microarray data downloaded and processed from GSE21510.
Some genes are missing:
"HSPA1A"   "HSPA1B"   "XCL2"     "HSPA8"    "NOTO"    
"HLA-DRB5" "HLA-DRB1" "KLRC2"    "KRT86" 
```{r, cache = T, warning=F, message=F, fig.cap="Expression of CD8 Res and Exh genes in the LCM  CRC data", fig.height=15, fig.width=8}
ll <- readRDS(paste0(ccleLcmPath, "ColoRectal_GeneExpression_LCM_MicroArray_GSE21510_AvergedProbes_Filtered.rds"))

gg <- selectedGenes[selectedGenes %in% rowData(ll)$Gene.Symbol]

l <- ll[rowData(ll)$Gene.Symbol %in% gg , ll$`tissue:ch1` == "cancer, LCM" ]

lAnnot <- colData(l)

lExpr <- assay(l)
row.names(lExpr) <- rowData(l)$Gene.Symbol

brbg <- rev(RColorBrewer::brewer.pal(11, "PRGn"))
brbg_hcl <- colorspace::diverging_hcl(11,
  h = c(180, 50), c = 80, l = c(20, 95), power = c(0.7, 1.3))

ComplexHeatmap::Heatmap(lExpr, 
                        col = brbg_hcl, 
                        show_column_names = F, name = "logExpr")

```


We also keep genes whose median expression is =< median of all genes in the data.
```{r}
th <- quantile(rowMedians(assay(ll)), probs = 0.5)
markersPassLCM <- rownames(lExpr)[rowMedians(lExpr) <= th]
markersPassLCM
#  [1] "NFKBID"   "RASGEF1B" "HAVCR2"   "PPP1R15A" "GNLY"     "BCL2A1"   "XCL1"     "LAG3"     "NR4A3"    "MYO7A"    "TWIST1"   "TNIP3"    "AFAP1L2" 
# [14] "TIGIT" 

# intersect(markersPassLCM, markersPassCCLE)
# [1] "RASGEF1B" "HAVCR2"   "GNLY"     "BCL2A1"   "XCL1"     "LAG3"     "NR4A3"    "MYO7A"    "TWIST1"   "TNIP3"    "TIGIT"    

# markersPassLCM[! markersPassLCM %in% markersPassCCLE]
# "NFKBID"   "PPP1R15A" "AFAP1L2"  
```



All Res and Exh genes that passed the threshold in either of the two data sets:
```{r}
passGenes <- unique(c(markersPassLCM, markersPassCCLE))
passGenes
```


Residency genes that pass the thresholds in either of the two data sets:
```{r}
ResPass <- passGenes[ passGenes %in% pctData_nkResPass$gene]
ResPass
```

Exhaustion genes that pass the thresholds in either of the two data sets:
```{r}
ExhPass <- passGenes[ passGenes %in% pctData_nkExhPass$gene]
ExhPass
```

Add these information to the signatures and export them.
```{r}

resData <- pctData_nkResPass

resData$CLsPass <- FALSE
resData$CLsPass[resData$gene %in% markersPassCCLE] <- TRUE

resData$LcmPass <- FALSE
resData$LcmPass[resData$gene %in% markersPassLCM] <- TRUE

resData$CancerPass <- FALSE
resData$CancerPass[resData$gene %in% ResPass] <- TRUE


exhData <- pctData_nkExhPass

exhData$CLsPass <- FALSE
exhData$CLsPass[exhData$gene %in% markersPassCCLE] <- TRUE

exhData$LcmPass <- FALSE
exhData$LcmPass[exhData$gene %in% markersPassLCM] <- TRUE

exhData$CancerPass <- FALSE
exhData$CancerPass[exhData$gene %in% ExhPass] <- TRUE

```


### Save merged markers
```{r}
exhData$Marker <- "Exh"
resData$Marker <- "Res"

# ResExhDataList <- list(exhData = exhData, 
#                        resData = resData)
# saveRDS(ResExhDataList, paste0(outPath, "ResExhMarkers_NK_CLs_LCM.RDS"))

ResExhData <- rbind(resData, exhData)

write.table(
  ResExhData,
  paste0(outPath, "ResExhMarkers_NK_CLs_LCM.txt"),
  row.names = F,
  sep = "\t"
)
```


# Session information
Here are the list of tools and packages (along with their versions) used in this document. 
```{r}
sessionInfo()
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 598px;"></div>

